---
title: "main"
output: html_document
date: "2026-03-24"
---

This is the main page of our project.

# We will start by loading the related libraries:

```{r}
library(TCGAbiolinks)
library(enrichplot)
library(ggplot2)
library(SummarizedExperiment)
library(tidyverse)
library(GEOquery)
library(DESeq2)
library(EnhancedVolcano)
library(clusterProfiler)
# this is the pathway library for a human (for GSEA)
library(org.Hs.eg.db)
library(msigdbr)
library(patchwork)
library(xCell2)
library(pheatmap)
```

# now let us access our data
```{r}
Bio.LIHC.query <- GDCquery(project = "TCGA-LIHC", 
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      experimental.strategy = "RNA-Seq")
Bio.LIHC.query.res <- getResults(Bio.LIHC.query)
```


# download
```{r}
options(timeout=1000)

GDCdownload(Bio.LIHC.query, directory = "./")
Bio.LIHC.data <- GDCprepare(Bio.LIHC.query, directory = "./")
```

# get LIHC metadata
```{r}
Bio.LIHC.metadata <- colData(Bio.LIHC.data)
Bio.LIHC.metadata_df <- as.data.frame(Bio.LIHC.metadata)
```

# check the data
```{r}
# NA values:
NAcolls <- colnames(Bio.LIHC.metadata_df)[sapply(Bio.LIHC.metadata_df, function(x) any(is.na(x)))]
# note that the definition of the sample has no NA so we will not clean anything for now

# amount of normal and cancerous samples:
normalSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "NT")
TumorSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "TP")

# total samples:
SampleCount <- length(Bio.LIHC.metadata_df$shortLetterCode)

# that doesnt add up!
notNormal_or_Tumor <- rownames(Bio.LIHC.metadata_df)[Bio.LIHC.metadata_df$shortLetterCode != "NT" & Bio.LIHC.metadata_df$shortLetterCode != "TP"]

Bio.LIHC.metadata_df[notNormal_or_Tumor, "shortLetterCode"]

# that is a reccurent tumor, which we don't want to deal with so i will remove them
Bio.LIHC.clean_metadata_df <- Bio.LIHC.metadata_df[Bio.LIHC.metadata_df$shortLetterCode != "TR",]

# how many NA are in the cancer stage?
cancerStageCountNA <- sum(is.na(Bio.LIHC.clean_metadata_df$ajcc_pathologic_stage))

# there are a lot of NA in there, so we will try some workarounds



```
# get the counts data
```{r}
counts.LIHC <- assay(Bio.LIHC.data)
clean_counts.LIHC <- counts.LIHC[,rownames(Bio.LIHC.clean_metadata_df)]

# check if everything is ready for DESeq
all(rownames(Bio.LIHC.clean_metadata_df) == colnames(clean_counts.LIHC))
sum(is.na(Bio.LIHC.clean_metadata_df$shortLetterCode))

```
# create DESeq object and perform DEseq
```{r}
LIHC.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC,
                              colData = Bio.LIHC.clean_metadata_df,
                              design = ~ shortLetterCode
)
LIHC.dds <- LIHC.dds[(rowSums(counts(LIHC.dds)) >= 10), ]
```

# define reference level
```{r}
LIHC.dds$shortLetterCode <- relevel(LIHC.dds$shortLetterCode, ref = "NT")
```
# run DESeq
```{r}
LIHC.dds <- DESeq(LIHC.dds)
resultsNames(LIHC.dds)
```
# get resaults
```{r}
res_TP_vs_NT.LFC <- lfcShrink(LIHC.dds, coef = "shortLetterCode_TP_vs_NT", type = "apeglm")

# clear the genes of their version number (important so GSEA will know what to do and so we can add the gene symbol)

rownames(res_TP_vs_NT.LFC) <- gsub("\\..*", "", rownames(res_TP_vs_NT.LFC))

# add symbols so we can "easily" interpret the genes
# Add gene symbols
res_TP_vs_NT.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_TP_vs_NT.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")
```
# perform pca to visualize 
```{r}
counts.vst.LIHC <- DESeq2::vst(LIHC.dds, blind = TRUE)

pcaData.LIHC <- DESeq2::plotPCA(counts.vst.LIHC, ntop = 1000, intgroup = c("shortLetterCode"), returnData = TRUE)
```

#visualize
```{r}
percentVar <- round(100 * attr(pcaData.LIHC, "percentVar"))
plot <- ggplot(pcaData.LIHC, aes(x=PC1, y=PC2, color=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))
plot
# looks awesome!
```

# now lets perform GSEA
# sort the genes by differentiallity
```{r}
# get differentiallity
oredered_genes.tumor_sample <- res_TP_vs_NT.LFC$log2FoldChange

# check for NA
sum(is.na(oredered_genes.tumor_sample))

# give names (cause thats just what you gotta do)
names(oredered_genes.tumor_sample) <- rownames(res_TP_vs_NT.LFC)

# order
oredered_genes.tumor_sample <- sort(oredered_genes.tumor_sample, decreasing = TRUE)



```
# look at the volcano plot
```{r fig.width=8, fig.height=8}

EnhancedVolcano::EnhancedVolcano(res_TP_vs_NT.LFC,
                lab = res_TP_vs_NT.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.00001,
                title = 'Tumor vs non tumor sample')

sort(res_TP_vs_NT.LFC$log2FoldChange, decreasing = TRUE)
```
lets look at some genes with high log_2 fold change and p value:
TERT - an important gene that helps in the production of telomerase which when activated, allows cell reproduction to continue (this is a highly expressed gene in most cancer, so a good sign)

CPLX2 - this is actually a cognitive related gene, that low expression of points to neorological decline (high expressivity seems not connected, but doesnt necciseraly mean anything)

LINC01419 - from what i read this gene literally just ruins everything and helps the cancer. in the nucleus it supresses tumor supressing genes. and it ruins somehow the DNA repair, which also helps tumor grow (one could say that the cancer is here because of the high expression, but im not going to take that jump)

# perform GSEA (general hallmark sets for a human)
```{r}
hallmark <- msigdbr(species = "Homo sapiens") %>% dplyr::select(gs_name, ensembl_gene)
gseaHallmark_res.tumor_sample <- GSEA(geneList = oredered_genes.tumor_sample,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)
```

# visualize
```{r}
dotplot(gseaHallmark_res.tumor_sample, showCategory = 10, split = ".sign", 
        title = "tumor sample enriched patways") +
   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```
# so there are really enriched and supressed genes (mostly supressed it seems). but we want to specifically look at the effect on immune pathways

```{r}
#hallmark_immune <- msigdbr(species = "Homo sapiens", category = "C7", subcategory = "IMMUNESIGDB") %>% dplyr::select(gs_name, ensembl_gene)
#gseaHallmark_immune_res.tumor_sample <- GSEA(geneList = oredered_genes.tumor_sample,
#                                      pvalueCutoff = 0.05,
#                                      nPermSimple = 10000,
#                                      TERM2GENE = hallmark_immune,
#                                      verbose = FALSE, 
#                                      eps = 0)
```
# visualize
```{r}

#dotplot(gseaHallmark_immune_res.tumor_sample, showCategory = 10, split=".sign", 
#        title = "tumor sample enriched patways") +
#   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```


```{r}
enrichplot_Inflammatory_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response")

enrichplot_Interferon_alpha_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response")

enrichplot_Interferon_gamma_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response")

enrichplot_TNFA_signaling <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling")


enrichplot_combined = enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = c("HALLMARK_INFLAMMATORY_RESPONSE",
                                    "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                    "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                    "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
                                    ),
                      title = "TNFA signaling")
enrichplot_Inflammatory_response
enrichplot_Interferon_alpha_response
enrichplot_Interferon_gamma_response
enrichplot_TNFA_signaling
```
all four pathways are an integral part of the immune system. When looking at the enrichment score graph, we can see that the genes are differentially suppressed in the tumor microenvironment. In this case, we say that the tumor is "cold" meaning, it suppresses immune activity.


# now to the xCell analysis
so we know that the immune system is suppressed around this tumor. but we want to know exactly what cells are suppressed. for that we use xCell2 and its immune reference.

```{r}
# first we get the normalized  counts from the differential expression
normalized_counts_matrix.LIHC <- counts(LIHC.dds, normalized = TRUE)
sum(is.na(rownames(normalized_counts_matrix.LIHC)))
# remove version number
rownames(normalized_counts_matrix.LIHC) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC))
# make symbol
gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC) <- gene_symbols[rownames(normalized_counts_matrix.LIHC)]
# no error! that's a relief. lets check for NA
sum(is.na(rownames(normalized_counts_matrix.LIHC)))
# thats a lot of na, so lets get them out
clean_normalized_counts_matrix.LIHC <- normalized_counts_matrix.LIHC[!is.na(rownames(normalized_counts_matrix.LIHC)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC)))
```
# start the actual analysis
```{r}
# load immune reference
data(ImmuneCompendium.xCell2Ref, package = "xCell2")

# run analysis
xCell_immune_res.LIHC <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC,
                               xcell2object = ImmuneCompendium.xCell2Ref)
```
# lets visualize the results
```{r fig.width=8, fig.height=8}
# we need an "annotation col" for some reason
annotation_col <- data.frame(Condition = Bio.LIHC.clean_metadata_df$shortLetterCode)
rownames(annotation_col) <- rownames(Bio.LIHC.clean_metadata_df)

# order the samples by sample type
ordered_for_heatmap <- rownames(Bio.LIHC.clean_metadata_df)[order(Bio.LIHC.clean_metadata_df$shortLetterCode)]
ordered_xCell2_res <- xCell_immune_res.LIHC[, ordered_for_heatmap]
pheatmap(ordered_xCell2_res, 
         annotation_col = annotation_col,
         scale = "row",
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         breaks = seq(-2, 2, length.out = 100),
         cutree_rows = 5,
         main = "xCell2 cell type heatmap")
```
we see a clear immune deregulation in the tumor samples compared to the non tumor samples.
especially in:
- granulocyte
- natural killer cell
- monocyte
- neutrophil
- plasmacytoid dendritic cell
- non classical monocyte
- eosinophil
- classical nonocyte
- basophil
and more with names too long.
by looking at the hierarcial clustering and cutoff, we can see that all supressed immune cells are closely related to each other. and by asking gemini (don't know what else i can use), these are all cells that are a big part of the "innate immune system" whos job is to be the "first line of defense" when there is some type of attack.
When the Innate Immune System is activated for too long, the Adaptive Immune System comes in. But with the Innate Immune System being active in normal samples and not tumor ones could point maybe towards an immune response, but not where the tumor is.