---
title: "Transcriptomic Signatures of Innate Immune Suppression and T-Cell Exhaustion in Hepatocellular Carcinoma"
author: "Re'em Melamed-Katz (213874373), Eliav Katzenstein (326986668)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "Figures/")
```

```{r libraries}
library(clusterProfiler)
library(data.table)
library(DESeq2)
library(EnhancedVolcano)
library(enrichplot)
library(GEOquery)
library(ggplot2)
library(ggpubr)
library(harmony)
library(Matrix)
library(msigdbr)
library(org.Hs.eg.db) # this is the pathway library for a human (for GSEA)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(Seurat)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(tidyverse)
library(xCell2)
```

***

## 1. Abstract (300 words in english)


***

## 2. Main analysis: Bulk RNA-seq

### 2.1 Prediction #1:

**Prediction**: 

**Biological reasoning**: 


### 2.2 Decision checkpoint #1:

**The decision**: 

**Why we chose it**: 

**Alternative considered**: 

**Why we rejected the alternative**: 


### 2.4 Code:

# now let us access our data
```{r}
Bio.LIHC.query <- GDCquery(project = "TCGA-LIHC", 
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      experimental.strategy = "RNA-Seq")
Bio.LIHC.query.res <- getResults(Bio.LIHC.query)
```


# download
```{r}
options(timeout=1000)

GDCdownload(Bio.LIHC.query, directory = "./")
Bio.LIHC.data <- GDCprepare(Bio.LIHC.query, directory = "./")
```

# get LIHC metadata
```{r}
Bio.LIHC.metadata <- colData(Bio.LIHC.data)
Bio.LIHC.metadata_df <- as.data.frame(Bio.LIHC.metadata)
```

# check the data
```{r}
# NA values:
NAcolls <- colnames(Bio.LIHC.metadata_df)[sapply(Bio.LIHC.metadata_df, function(x) any(is.na(x)))]
# note that the definition of the sample has no NA so we will not clean anything for now

# amount of normal and cancerous samples:
normalSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "NT")
TumorSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "TP")

# total samples:
SampleCount <- length(Bio.LIHC.metadata_df$shortLetterCode)

# that doesnt add up!
notNormal_or_Tumor <- rownames(Bio.LIHC.metadata_df)[Bio.LIHC.metadata_df$shortLetterCode != "NT" & Bio.LIHC.metadata_df$shortLetterCode != "TP"]

Bio.LIHC.metadata_df[notNormal_or_Tumor, "shortLetterCode"]

# that is a reccurent tumor, which we don't want to deal with so i will remove them
Bio.LIHC.clean_metadata_df <- Bio.LIHC.metadata_df[Bio.LIHC.metadata_df$shortLetterCode != "TR",]

# how many NA are in the cancer stage?
cancerStageCountNA <- sum(is.na(Bio.LIHC.clean_metadata_df$ajcc_pathologic_stage))

# there are a lot of NA in there, so we will try some workarounds



```

# get the counts data
```{r}
counts.LIHC <- assay(Bio.LIHC.data)
clean_counts.LIHC <- counts.LIHC[,rownames(Bio.LIHC.clean_metadata_df)]

# check if everything is ready for DESeq
all(rownames(Bio.LIHC.clean_metadata_df) == colnames(clean_counts.LIHC))
sum(is.na(Bio.LIHC.clean_metadata_df$shortLetterCode))

```

# create DESeq object and perform DEseq
```{r}
LIHC.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC,
                              colData = Bio.LIHC.clean_metadata_df,
                              design = ~ shortLetterCode
)
LIHC.dds <- LIHC.dds[(rowSums(counts(LIHC.dds)) >= 10), ]
```

# define reference level
```{r}
LIHC.dds$shortLetterCode <- relevel(LIHC.dds$shortLetterCode, ref = "NT")
```

# run DESeq
```{r}
LIHC.dds <- DESeq(LIHC.dds)
resultsNames(LIHC.dds)
```

# get resaults
```{r}
res_TP_vs_NT.LFC <- lfcShrink(LIHC.dds, coef = "shortLetterCode_TP_vs_NT", type = "apeglm")

# clear the genes of their version number (important so GSEA will know what to do and so we can add the gene symbol)

rownames(res_TP_vs_NT.LFC) <- gsub("\\..*", "", rownames(res_TP_vs_NT.LFC))

# add symbols so we can "easily" interpret the genes
# Add gene symbols
res_TP_vs_NT.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_TP_vs_NT.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")
```

# perform pca to visualize 
```{r}
counts.vst.LIHC <- DESeq2::vst(LIHC.dds, blind = TRUE)

pcaData.LIHC <- DESeq2::plotPCA(counts.vst.LIHC, ntop = 1000, intgroup = c("shortLetterCode"), returnData = TRUE)
```

# visualize
```{r}
percentVar <- round(100 * attr(pcaData.LIHC, "percentVar"))
plot <- ggplot(pcaData.LIHC, aes(x=PC1, y=PC2, color=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))
plot
# looks awesome!
```

# now lets perform GSEA
# sort the genes by differentiallity
```{r}
# get differentiallity
oredered_genes.tumor_sample <- res_TP_vs_NT.LFC$log2FoldChange

# check for NA
sum(is.na(oredered_genes.tumor_sample))

# give names (cause thats just what you gotta do)
names(oredered_genes.tumor_sample) <- rownames(res_TP_vs_NT.LFC)

# order
oredered_genes.tumor_sample <- sort(oredered_genes.tumor_sample, decreasing = TRUE)



```
# look at the volcano plot
```{r fig.width=8, fig.height=8}

EnhancedVolcano::EnhancedVolcano(res_TP_vs_NT.LFC,
                lab = res_TP_vs_NT.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.00001,
                title = 'Tumor vs non tumor sample')

sort(res_TP_vs_NT.LFC$log2FoldChange, decreasing = TRUE)
```
lets look at some genes with high log_2 fold change and p value:
TERT - an important gene that helps in the production of telomerase which when activated, allows cell reproduction to continue (this is a highly expressed gene in most cancer, so a good sign)

CPLX2 - this is actually a cognitive related gene, that low expression of points to neorological decline (high expressivity seems not connected, but doesnt necciseraly mean anything)

LINC01419 - from what i read this gene literally just ruins everything and helps the cancer. in the nucleus it supresses tumor supressing genes. and it ruins somehow the DNA repair, which also helps tumor grow (one could say that the cancer is here because of the high expression, but im not going to take that jump)

# perform GSEA (general hallmark sets for a human)
```{r}
hallmark <- msigdbr(species = "Homo sapiens") %>% dplyr::select(gs_name, ensembl_gene)
gseaHallmark_res.tumor_sample <- GSEA(geneList = oredered_genes.tumor_sample,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)
```

# visualize
```{r}
dotplot(gseaHallmark_res.tumor_sample, showCategory = 10, split = ".sign", 
        title = "tumor sample enriched patways") +
   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```
# so there are really enriched and supressed genes (mostly supressed it seems). but we want to specifically look at the effect on immune pathways

```{r}
#hallmark_immune <- msigdbr(species = "Homo sapiens", category = "C7", subcategory = "IMMUNESIGDB") %>% dplyr::select(gs_name, ensembl_gene)
#gseaHallmark_immune_res.tumor_sample <- GSEA(geneList = oredered_genes.tumor_sample,
#                                      pvalueCutoff = 0.05,
#                                      nPermSimple = 10000,
#                                      TERM2GENE = hallmark_immune,
#                                      verbose = FALSE, 
#                                      eps = 0)
```

# visualize
```{r}

#dotplot(gseaHallmark_immune_res.tumor_sample, showCategory = 10, split=".sign", 
#        title = "tumor sample enriched patways") +
#   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```


```{r}
enrichplot_Inflammatory_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response")

enrichplot_Interferon_alpha_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response")

enrichplot_Interferon_gamma_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response")

enrichplot_TNFA_signaling <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling")


enrichplot_combined = enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = c("HALLMARK_INFLAMMATORY_RESPONSE",
                                    "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                    "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                    "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
                                    ),
                      title = "TNFA signaling")
enrichplot_Inflammatory_response
enrichplot_Interferon_alpha_response
enrichplot_Interferon_gamma_response
enrichplot_TNFA_signaling
```
all four pathways are an integral part of the immune system. When looking at the enrichment score graph, we can see that the genes are differentially suppressed in the tumor microenvironment. In this case, we say that the tumor is "cold" meaning, it suppresses immune activity.


# now to the xCell analysis
so we know that the immune system is suppressed around this tumor. but we want to know exactly what cells are suppressed. for that we use xCell2 and its immune reference.

```{r}
# first we get the normalized  counts from the differential expression
normalized_counts_matrix.LIHC <- counts(LIHC.dds, normalized = TRUE)
sum(is.na(rownames(normalized_counts_matrix.LIHC)))
# remove version number
rownames(normalized_counts_matrix.LIHC) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC))
# make symbol
gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC) <- gene_symbols[rownames(normalized_counts_matrix.LIHC)]
# no error! that's a relief. lets check for NA
sum(is.na(rownames(normalized_counts_matrix.LIHC)))
# thats a lot of na, so lets get them out
clean_normalized_counts_matrix.LIHC <- normalized_counts_matrix.LIHC[!is.na(rownames(normalized_counts_matrix.LIHC)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC)))
```

# start the actual analysis
```{r}
# load immune reference
data(ImmuneCompendium.xCell2Ref, package = "xCell2")

# run analysis
xCell_immune_res.LIHC <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC,
                               xcell2object = ImmuneCompendium.xCell2Ref)
```

# lets visualize the results
```{r fig.width=8, fig.height=8}
# we need an "annotation col" for some reason
annotation_col <- data.frame(Condition = Bio.LIHC.clean_metadata_df$shortLetterCode)
rownames(annotation_col) <- rownames(Bio.LIHC.clean_metadata_df)

# order the samples by sample type
ordered_for_heatmap <- rownames(Bio.LIHC.clean_metadata_df)[order(Bio.LIHC.clean_metadata_df$shortLetterCode)]
ordered_xCell2_res <- xCell_immune_res.LIHC[, ordered_for_heatmap]
pheatmap(ordered_xCell2_res, 
         annotation_col = annotation_col,
         scale = "row",
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         breaks = seq(-2, 2, length.out = 100),
         cutree_rows = 5,
         main = "xCell2 cell type heatmap")
```

we see a clear immune deregulation in the tumor samples compared to the non tumor samples.
especially in:
- granulocyte
- natural killer cell
- monocyte
- neutrophil
- plasmacytoid dendritic cell
- non classical monocyte
- eosinophil
- classical nonocyte
- basophil
and more with names too long.
by looking at the hierarcial clustering and cutoff, we can see that all supressed immune cells are closely related to each other. and by asking gemini (don't know what else i can use), these are all cells that are a big part of the "innate immune system" whos job is to be the "first line of defense" when there is some type of attack.
When the Innate Immune System is activated for too long, the Adaptive Immune System comes in. But with the Innate Immune System being active in normal samples and not tumor ones could point maybe towards an immune response, but not where the tumor is.


### 2.5 Results:

**Result comparison (prediction #1)**: 

**Biological Integration**:


***

## 3. Secondary analysis: scRNA-Seq

### 3.1 Prediction #2

**Prediction**: 

**Biological reasoning**: 


### 2.3 Decision checkpoint #2:

**The decision**: 

**Why we chose it**: 

**Alternative considered**: 

**Why we rejected the alternative**:


### 3.2 Code:

```{r harmony}
# load 1st dataset
sc_data_1 <- ReadMtx(
  mtx = "Data/scRNA_GSE140228/matrix.mtx",
  cells = "Data/scRNA_GSE140228/barcodes.tsv",
  features = "Data/scRNA_GSE140228/features.tsv",
  skip.feature = 1
)
seu_1 <- CreateSeuratObject(counts = sc_data_1, project = "GSE140228", min.cells = 3, min.features = 200)

# load 2nd dataset
file_path_2 <- "Data/scRNA_GSE149614/GSE149614_HCC.scRNAseq.S71915.count.txt"
sc_data_2_raw <- fread(file_path_2, data.table = FALSE, select = 1:3001)
rownames(sc_data_2_raw) <- sc_data_2_raw[, 1]
sc_data_2_clean <- sc_data_2_raw[, -1]
sc_data_2_sparse <- as(as.matrix(sc_data_2_clean), "dgCMatrix")

# clean up memory
rm(sc_data_2_raw, sc_data_2_clean)
gc()

seu_2 <- CreateSeuratObject(counts = sc_data_2_sparse, project = "GSE149614", min.cells = 3, min.features = 200)

# merge and filter out low quality cells
seu_merged <- merge(seu_1, y = seu_2, add.cell.ids = c("DS1", "DS2"), project = "HCC_Integrated")
seu_merged[["percent.mt"]] <- PercentageFeatureSet(seu_merged, pattern = "^MT-")
seu_merged <- subset(seu_merged, subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mt < 10)

# pipeline and Harmony batch correction
seu_merged <- NormalizeData(seu_merged) %>% 
              FindVariableFeatures() %>% 
              ScaleData() %>% 
              RunPCA() %>% 
              RunHarmony(group.by.vars = "orig.ident")

# clustering and UMAP
seu_merged <- FindNeighbors(seu_merged, reduction = "harmony", dims = 1:15)
seu_merged <- FindClusters(seu_merged, resolution = 0.5)
seu_merged <- RunUMAP(seu_merged, reduction = "harmony", dims = 1:15)

# visual check plots
DimPlot(seu_merged, reduction = "umap", label = TRUE) + ggtitle("Cluster Map")
FeaturePlot(seu_merged, features = "CD8A") + ggtitle("CD8+ T-cell Localization")
```

We examined the UMAP and CD8A FeaturePlot to localize the T-cell compartment, and identified clusters 0, 1, 6, 14, 17 as our target populations.


### 3.2 Decision checkpoint #3: cluster characterization method

**The decision**: We chose to use logistic regression with dataset integration.

**Why we chose it**: Although we integrated our datasets using Harmony, Harmony only corrects the low dimensional embedding (PCA and UMAP) to allow for shared clustering. It does not modify the underlying gene expression matrix. By using logistic regression with latent.vars = "orig.ident", we can explicitly account for and neutralize the batch effects (technical variation between the two studies) during the actual marker discovery. This ensures our markers are biologically driven by cell type and not artifacts of different laboratories.

**Alternative considered**: Wilcoxon rank-sum test.

**Why we rejected the alternative**: The Wilcoxon test is a non-parametric test that compares ranks but cannot incorporate covariates. Using it on merged datasets without accounting for batch effects often leads to "false positive" markers that represent dataset differences rather than true cell-type biology.


### 3.3 Code:
```{r lr}
target_clusters <- c(0, 1, 6, 14, 17)

# join all layers
seu_merged <- JoinLayers(seu_merged)

# extract our T-cell group markers
t_markers <- FindMarkers(seu_merged, 
                         ident.1 = target_clusters, 
                         test.use = "LR", 
                         latent.vars = "orig.ident", 
                         only.pos = TRUE, 
                         min.pct = 0.25)

# keep only the significant genes
sig_genes <- rownames(t_markers[t_markers$p_val_adj < 0.05 & t_markers$avg_log2FC > 0.5, ])

# GO
ego_results <- enrichGO(gene          = sig_genes,
                        OrgDb         = org.Hs.eg.db,
                        keyType       = 'SYMBOL',
                        ont           = "BP", 
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

dotplot(ego_results, showCategory = 15, label_format = 40) + 
  theme_bw() + 
  scale_color_gradient(low = "red", high = "blue") +
  ggtitle("Functional Enrichment: CD8+ T-cell Compartment") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

Now, we will examine if our target T-cells actively co-express classical exhaustion markers. 
We will also test if this overall exhaustion signature is significantly stronger in the tumor environment compared to normal tissue:

```{r results}
exhaustion_list <- list(c("PDCD1", "LAG3", "HAVCR2", "TIGIT", "CTLA4", "TOX", "LAYN"))

seu_merged <- AddModuleScore(
  object = seu_merged,
  features = exhaustion_list,
  name = "T_Cell_Exhaustion_Score"
)

exhaustion_markers <- c("CD8A", "PDCD1", "LAG3", "HAVCR2", "TIGIT", "TOX")

p1 <- DotPlot(seu_merged, features = exhaustion_markers, idents = target_clusters) +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu"))) +
  labs(title = "Marker Expression in T-cell Clusters", x = "Genes", y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- FeaturePlot(seu_merged, features = "T_Cell_Exhaustion_Score1") +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu"))) +
  ggtitle("Global T-Cell Exhaustion Score")

meta_ds1 <- read.delim("Data/scRNA_GSE140228/GSE140228_UMI_counts_Droplet_cellinfo.tsv", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds1) <- paste0("DS1_", meta_ds1$Barcode)
condition_ds1 <- meta_ds1[, "Tissue", drop = FALSE]
colnames(condition_ds1) <- "Condition"

meta_ds2 <- read.delim("Data/scRNA_GSE149614/GSE149614_HCC.metadata.updated.txt", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds2) <- paste0("DS2_", meta_ds2$Cell)
condition_ds2 <- meta_ds2[, "site", drop = FALSE]
colnames(condition_ds2) <- "Condition"

combined_meta <- rbind(condition_ds1, condition_ds2)
seu_merged <- AddMetaData(seu_merged, metadata = combined_meta)

seu_merged$Condition[seu_merged$Condition == "Adjacent"] <- "Normal"

seu_filtered <- subset(seu_merged, subset = Condition %in% c("Tumor", "Normal"))

p3 <- VlnPlot(seu_filtered, features = "T_Cell_Exhaustion_Score1", 
              idents = target_clusters, group.by = "Condition", pt.size = 0) +
  ggtitle("Exhaustion Score: Normal vs Tumor") +
  stat_compare_means(method = "wilcox.test")

print(p1)
print(p2)
print(p3)
```

### 3.4 Results:

**Result comparison (prediction #2)**: 

**Biological Integration**:
