---
title: "Transcriptomic Signatures of Innate Immune Suppression and T-Cell Exhaustion in Hepatocellular Carcinoma"
author: "Re'em Melamed-Katz (213874373), Eliav Katzenstein (326986668)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "Figures/")
```

```{r libraries}
library(clusterProfiler)
library(data.table)
library(DESeq2)
library(EnhancedVolcano)
library(enrichplot)
library(GEOquery)
library(ggplot2)
library(ggpubr)
library(harmony)
library(Matrix)
library(msigdbr)
library(org.Hs.eg.db) # this is the pathway library for a human (for GSEA)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(Seurat)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(tidyverse)
library(xCell2)
```

***

## 1. Abstract (300 words in english)

***

## 2. Main analysis: Bulk RNA-seq

### Data Acquisition and Preprocessing

**Decision checkpoint #1: Dataset Selection**

* **The decision**: TCGA-LIHC as the dataset.
* **Why we chose it**: It is crucial, seeing as it is very general and comprehensive with many samples (though only with 50 normal samples).
* **Alternative considered**: GSE-144269.
* **Why we rejected the alternative**: Seeing as it was smaller and only had patients with prior HepD that interfered with the results.

This is actually the first decision checkpoint. The choice of TCGA-LIHC
as the dataset is crucial, seeing as it is very general and
comprehensive with many samples (though only with 50 normal samples). An
alternative we had was GSE-144269. But seeing as it was smaller and only
had patients with prior HepD that interfered with the results.

```{r}
# get the data
Bio.LIHC.query <- GDCquery(project = "TCGA-LIHC", 
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      experimental.strategy = "RNA-Seq")
Bio.LIHC.query.res <- getResults(Bio.LIHC.query)
```

```{r}
# download
options(timeout=1000)

GDCdownload(Bio.LIHC.query, directory = "./")
Bio.LIHC.data <- GDCprepare(Bio.LIHC.query, directory = "./")
```

```{r}
# get the LIHC metadata
Bio.LIHC.metadata <- colData(Bio.LIHC.data)
Bio.LIHC.metadata_df <- as.data.frame(Bio.LIHC.metadata)
```

```{r}
# check the data to make sure its clean

# NA values:
NAcolls <- colnames(Bio.LIHC.metadata_df)[sapply(Bio.LIHC.metadata_df, function(x) any(is.na(x)))]
# note that the definition of the sample has no NA so we will not clean anything for now

# amount of normal and cancerous samples:
normalSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "NT")
TumorSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "TP")

# total samples:
SampleCount <- length(Bio.LIHC.metadata_df$shortLetterCode)

# that doesnt add up!
notNormal_or_Tumor <- rownames(Bio.LIHC.metadata_df)[Bio.LIHC.metadata_df$shortLetterCode != "NT" & Bio.LIHC.metadata_df$shortLetterCode != "TP"]

Bio.LIHC.metadata_df[notNormal_or_Tumor, "shortLetterCode"]

# that is a recurrent tumor, which we don't want to deal with so i will remove them
Bio.LIHC.clean_metadata_df <- Bio.LIHC.metadata_df[Bio.LIHC.metadata_df$shortLetterCode != "TR",]

# how many NA are in the cancer stage?
cancerStageCountNA <- sum(is.na(Bio.LIHC.clean_metadata_df$ajcc_pathologic_stage))

```

```{r}
# get the Seq counts data

counts.LIHC <- assay(Bio.LIHC.data)
clean_counts.LIHC <- counts.LIHC[,rownames(Bio.LIHC.clean_metadata_df)]

# check if everything is ready for DESeq
all(rownames(Bio.LIHC.clean_metadata_df) == colnames(clean_counts.LIHC))
sum(is.na(Bio.LIHC.clean_metadata_df$shortLetterCode))

```


### Differential Expression Analysis

**Decision checkpoint #2: Differential Expression Tool**

* **The decision**: We decided to use DESeq2 as our differential expression tool.
* **Why we chose it**: It is the one we were most familiar with, a standard and reliable tool.
* **Alternative considered**: edgeR.
* **Why we rejected the alternative**: Preferred DESeq2 for familiarity.

```{r}
# create DESeq object and perform DEseq
LIHC.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC,
                              colData = Bio.LIHC.clean_metadata_df,
                              design = ~ shortLetterCode
)
LIHC.dds <- LIHC.dds[(rowSums(counts(LIHC.dds)) >= 10), ]
```

**define reference level**

```{r}
LIHC.dds$shortLetterCode <- relevel(LIHC.dds$shortLetterCode, ref = "NT")
```

**run DESeq**

```{r}
LIHC.dds <- DESeq(LIHC.dds)
resultsNames(LIHC.dds)
```

**get results**

```{r}
res_TP_vs_NT.LFC <- lfcShrink(LIHC.dds, coef = "shortLetterCode_TP_vs_NT", type = "apeglm")

# clear the genes of their version number (important so GSEA will know what to do and so we can add the gene symbol)

rownames(res_TP_vs_NT.LFC) <- gsub("\\..*", "", rownames(res_TP_vs_NT.LFC))

# add symbols so we can "easily" interpret the genes
# Add gene symbols
res_TP_vs_NT.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_TP_vs_NT.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")
```

**look at the volcano plot**

```{r fig.width=8, fig.height=8}

EnhancedVolcano::EnhancedVolcano(res_TP_vs_NT.LFC,
                lab = res_TP_vs_NT.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.00001,
                title = 'Tumor vs non tumor sample')

sort(res_TP_vs_NT.LFC$log2FoldChange, decreasing = TRUE)
```

lets look at some genes with high log_2 fold change and p value: TERT -
an important gene that helps in the production of telomerase which when
activated, allows cell reproduction to continue (this is a highly
expressed gene in most cancer, so a good sign)

CPLX2 - this is actually a cognitive related gene, that low expression
of points to neorological decline (high expressivity seems not
connected, but doesnt necciseraly mean anything)

LINC01419 - from what i read this gene literally just ruins everything
and helps the cancer. in the nucleus it supresses tumor supressing
genes. and it ruins somehow the DNA repair, which also helps tumor grow
(one could say that the cancer is here because of the high expression,
but im not going to take that jump)

**perform pca to visualize**

```{r}
counts.vst.LIHC <- DESeq2::vst(LIHC.dds, blind = TRUE)

pcaData.LIHC <- DESeq2::plotPCA(counts.vst.LIHC, ntop = 1000, intgroup = c("shortLetterCode"), returnData = TRUE)
```

**visualize**

```{r}
percentVar <- round(100 * attr(pcaData.LIHC, "percentVar"))
plot <- ggplot(pcaData.LIHC, aes(x=PC1, y=PC2, color=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))
plot
# looks awesome!
```

### Pathway Enrichment Analysis

**Decision checkpoint #3: Pathway Analysis Tool**

* **The decision**: GSEA is an excellent tool for the job.
* **Why we chose it**: Seeing as we want to check the immune pathways (as in, know a lot about a little). By choosing specific pathways (as seen later), we can see excatly how and how much is that pathway differentially expressed.
* **Alternative considered**: ORA.
* **Why we rejected the alternative**: It is much less sophisticated and robust, and it is much more suited for quick small answers, while we need this as a robust and main part of our analysis.

Seeing as we want to check the immune pathways (as in, know a lot about
a little), GSEA is an excellent tool for the job. By choosing specific
pathways (as seen later), we can see excatly how and how much is that
pathway differentially expressed. An alternative was ORA, but it is much
less sophisticated and robust, and it is much more suited for quick
small answers, while we need this as a robust and main part of our
analysis.

```{r}
# GSEA
# sort the genes by differentiallity

# get differentiallity
oredered_genes.tumor_sample <- res_TP_vs_NT.LFC$log2FoldChange

# check for NA
sum(is.na(oredered_genes.tumor_sample))

# give names (cause thats just what you gotta do)
names(oredered_genes.tumor_sample) <- rownames(res_TP_vs_NT.LFC)

# order
oredered_genes.tumor_sample <- sort(oredered_genes.tumor_sample, decreasing = TRUE)



```

A p-value of 0.05 was chosen as a standard p-value for statistical significance.

```{r}
# perform GSEA (using hallmark sets for a human)
hallmark <- msigdbr(species = "Homo sapiens") %>% dplyr::select(gs_name, ensembl_gene)
gseaHallmark_res.tumor_sample <- GSEA(geneList = oredered_genes.tumor_sample,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)
```

```{r}
# visualize

dotplot(gseaHallmark_res.tumor_sample, showCategory = 10, split = ".sign", 
        title = "tumor sample enriched patways") +
   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```

so there are really enriched and supressed genes (mostly supressed it
seems). but we want to specifically look at the effect on immune
pathways

```{r}
#hallmark_immune <- msigdbr(species = "Homo sapiens", category = "C7", subcategory = "IMMUNESIGDB") %>% dplyr::select(gs_name, ensembl_gene)
#gseaHallmark_immune_res.tumor_sample <- GSEA(geneList = oredered_genes.tumor_sample,
#                                      pvalueCutoff = 0.05,
#                                      nPermSimple = 10000,
#                                      TERM2GENE = hallmark_immune,
#                                      verbose = FALSE, 
#                                      eps = 0)
```

**visualize**

```{r}

#dotplot(gseaHallmark_immune_res.tumor_sample, showCategory = 10, split=".sign", 
#        title = "tumor sample enriched patways") +
#   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```


**Prediction #1: Immune Pathway Activity**

* **Prediction**: The enrichment score of the 4 selected immune pathways can go both ways. We would assume that there will be variance, meaning the immune pathways are either supressed or activated.
* **Biological reasoning**: As learned during the tutorials, this will point towards the tumor being "hot" or "cold" meaning an active immune environment and a supressed immune environment respectively.

```{r}
enrichplot_Inflammatory_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response")

enrichplot_Interferon_alpha_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response")

enrichplot_Interferon_gamma_response <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response")

enrichplot_TNFA_signaling <- enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling")


enrichplot_combined = enrichplot::gseaplot2(gseaHallmark_res.tumor_sample, 
                      geneSetID = c("HALLMARK_INFLAMMATORY_RESPONSE",
                                    "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                    "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                    "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
                                    ),
                      title = "TNFA signaling")
enrichplot_Inflammatory_response
enrichplot_Interferon_alpha_response
enrichplot_Interferon_gamma_response
enrichplot_TNFA_signaling
```

all four pathways are an integral part of the immune system. When
looking at the enrichment score graph, we can see that the genes are
differentially suppressed in the tumor microenvironment. In this case,
we say that the tumor is "cold" meaning, it suppresses immune activity.

### Tumor Microenvironment Deconvolution (xCell)

**Prediction #2: Immune Cell Suppression**

* **Prediction**: Seeing as the 4 immune pathways we chose are suppressed around the tumor, we would assume that all if not some of the immune cells connected to that pathway would be supressed.
Inflammatory Response: "Macrophages", "Neutrophils", "T-Cells", "B-Cells", "NK-Cells", "D-Cells"
Interfron Alpha Response: "NK-Cells", "B-Cells", "T-Cells", "D-Cells"
Interfron Gamma Response: "Macrophages", "T Helper Cells", "CD8+ Cytotoxic T-Cells", "NK-Cells", "B-Cells", "D-Cells"
TNFA signaling: "Macrophages", "T-Cells", "B-Cells", "NK-Cells", "Fibroblasts", "Keratinocytes"
* **Biological reasoning**: We know that the immune system is suppressed around this tumor, but we want to know exactly what cells are suppressed. For that we map the pathways to their cell types using xCell2 and its immune reference.

**xCell analysis**

so we know that the immune system is suppressed around this tumor. but
we want to know exactly what cells are suppressed. for that we use
xCell2 and its immune reference.

```{r}
# first we get the normalized  counts from the differential expression
normalized_counts_matrix.LIHC <- counts(LIHC.dds, normalized = TRUE)
sum(is.na(rownames(normalized_counts_matrix.LIHC)))
# remove version number
rownames(normalized_counts_matrix.LIHC) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC))
# make symbol
gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC) <- gene_symbols[rownames(normalized_counts_matrix.LIHC)]
# no error! that's a relief. lets check for NA
sum(is.na(rownames(normalized_counts_matrix.LIHC)))
# thats a lot of na, so lets get them out
clean_normalized_counts_matrix.LIHC <- normalized_counts_matrix.LIHC[!is.na(rownames(normalized_counts_matrix.LIHC)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC)))
```

```{r}
# start the actual analysis

# load immune reference
data(ImmuneCompendium.xCell2Ref, package = "xCell2")

# run analysis
xCell_immune_res.LIHC <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC,
                               xcell2object = ImmuneCompendium.xCell2Ref)
```

```{r fig.width=8, fig.height=8}
# lets visualize the results

# we need an "annotation col" for some reason
annotation_col <- data.frame(Condition = Bio.LIHC.clean_metadata_df$shortLetterCode)
rownames(annotation_col) <- rownames(Bio.LIHC.clean_metadata_df)

# order the samples by sample type
ordered_for_heatmap <- rownames(Bio.LIHC.clean_metadata_df)[order(Bio.LIHC.clean_metadata_df$shortLetterCode)]
ordered_xCell2_res <- xCell_immune_res.LIHC[, ordered_for_heatmap]
pheatmap(ordered_xCell2_res, 
         annotation_col = annotation_col,
         scale = "row",
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         breaks = seq(-2, 2, length.out = 100),
         cutree_rows = 5,
         main = "xCell2 cell type heatmap")
```


### Results: Bulk RNA-seq

* **Result comparison (prediction #1 & #2)**: we see a clear immune deactivation in the tumor samples compared to the non tumor samples, when looking at: granulocyte, natural killer cell, monocyte, neutrophil, plasmacytoid dendritic cell, non classical monocyte, eosinophil, classical nonocyte, basophil and more with names too long.
* **Biological Integration**: by looking at the hierarcial clustering and cutoff, we can see that all supressed immune cells are closely related to each other. and by asking gemini (don't know what else i can use), these are all cells that are a big part of the "innate immune system" who's job is to be the "first line of defense" when there is some type of attack. When the Innate Immune System is activated for too long, the Adaptive Immune System comes in. But with the Innate Immune System being active in normal samples and not tumor ones could point maybe towards an immune response, but not where the tumor is.

we see a clear immune deactivation in the tumor samples compared to the
non tumor samples, when looking at: - granulocyte - natural killer cell -
monocyte - neutrophil - plasmacytoid dendritic cell - non classical
monocyte - eosinophil - classical nonocyte - basophil and more with
names too long. 
by looking at the hierarcial clustering and cutoff, we
can see that all supressed immune cells are closely related to each
other. and by asking gemini (don't know what else i can use), these are
all cells that are a big part of the "innate immune system" who's job is
to be the "first line of defense" when there is some type of attack.
When the Innate Immune System is activated for too long, the Adaptive
Immune System comes in. But with the Innate Immune System being active
in normal samples and not tumor ones could point maybe towards an immune
response, but not where the tumor is.

***

## 3. Secondary analysis: scRNA-Seq

**Prediction #3: T-cell Exhaustion**

* **Prediction**: CD8+ T-cells infiltrating the tumor microenvironment will express a high level of T-cell exhaustion markers compared to T-cells in normal adjacent tissue.
* **Biological reasoning**: In cancer, continuous exposure to tumor antigens causes T-cells to enter a state of exhaustion. As T-cells become exhausted, they lose their ability to kill target cells and upregulate multiple inhibitory receptors, including PD-1 (PDCD1), LAG-3, and TIGIT. Since hepatocellular carcinoma provides a constant antigen source, we expect CD8+ T-cells within the tumor to co-express these exhaustion markers compared to cells in the healthy adjacent tissue.

### Data Processing and Integration

```{r harmony}
# load 1st dataset
sc_data_1 <- ReadMtx(
  mtx = "Data/scRNA_GSE140228/matrix.mtx",
  cells = "Data/scRNA_GSE140228/barcodes.tsv",
  features = "Data/scRNA_GSE140228/features.tsv",
  skip.feature = 1
)
seu_1 <- CreateSeuratObject(counts = sc_data_1, project = "GSE140228", min.cells = 3, min.features = 200)

rm(sc_data_1)
gc()

# load 2nd dataset
file_path_2 <- "Data/scRNA_GSE149614/GSE149614_HCC.scRNAseq.S71915.count.txt"
sc_data_2_raw <- fread(file_path_2, data.table = FALSE, select = 1:3001)
rownames(sc_data_2_raw) <- sc_data_2_raw[, 1]
sc_data_2_clean <- sc_data_2_raw[, -1]

rm(sc_data_2_raw) 
gc()

sc_data_2_matrix <- as.matrix(sc_data_2_clean)
rm(sc_data_2_clean)
gc()

sc_data_2_sparse <- as(sc_data_2_matrix, "dgCMatrix")
rm(sc_data_2_matrix)
gc()

seu_2 <- CreateSeuratObject(counts = sc_data_2_sparse, project = "GSE149614", min.cells = 3, min.features = 200)

rm(sc_data_2_sparse) 
gc()

# merge and filter out low quality cells
seu_merged <- merge(seu_1, y = seu_2, add.cell.ids = c("DS1", "DS2"), project = "HCC_Integrated")

rm(seu_1, seu_2)
gc()

seu_merged[["percent.mt"]] <- PercentageFeatureSet(seu_merged, pattern = "^MT-")
seu_merged <- subset(seu_merged, subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mt < 10)

# pipeline and Harmony batch correction
seu_merged <- NormalizeData(seu_merged) %>% 
              FindVariableFeatures() %>% 
              ScaleData() %>% 
              RunPCA() %>% 
              RunHarmony(group.by.vars = "orig.ident")

# clustering and UMAP
seu_merged <- FindNeighbors(seu_merged, reduction = "harmony", dims = 1:15)
seu_merged <- FindClusters(seu_merged, resolution = 0.5)
seu_merged <- RunUMAP(seu_merged, reduction = "harmony", dims = 1:15)

# visual check plots
DimPlot(seu_merged, reduction = "umap", label = TRUE) + ggtitle("Cluster Map")
FeaturePlot(seu_merged, features = "CD8A") + ggtitle("CD8+ T-cell Localization")

```

We examined the UMAP and CD8A FeaturePlot to localize the T-cell compartment, and identified clusters 0, 1, 6, 14, 17 as our target populations.

### Cluster Characterization and Analysis

**Decision checkpoint #4: cluster characterization method**

* **The decision**: We calculated an aggregated module score (using Seurat's AddModuleScore) based on a predefined list of exhaustion genes instead of analyzing individual markers.
* **Why we chose it**: Single-cell RNA-seq data contains many technical dropouts. Aggregating the signal from multiple exhaustion markers creates a more stable cell-level metric, reducing technical noise.
* **Alternative considered**: Comparing the expression of a single exhaustion marker like PDCD1 between normal and tumor cells.
* **Why we rejected the alternative**: Looking at a single gene in scRNA-seq data leads to high false-negative rates. Many true exhausted T-cells might have zero counts for PDCD1 due to sequencing limitations, which would hurt our statistical power.

```{r results}
target_clusters <- c(0, 1, 6, 14, 17)

seu_merged <- subset(seu_merged, downsample = 500)

# join all layers
seu_merged <- JoinLayers(seu_merged)

exhaustion_list <- list(c("PDCD1", "LAG3", "HAVCR2", "TIGIT", "CTLA4", "TOX", "LAYN"))

seu_merged <- AddModuleScore(
  object = seu_merged,
  features = exhaustion_list,
  name = "T_Cell_Exhaustion_Score"
)

exhaustion_markers <- c("CD8A", "PDCD1", "LAG3", "HAVCR2", "TIGIT", "TOX")

p1 <- DotPlot(seu_merged, features = exhaustion_markers, idents = target_clusters) +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu"))) +
  labs(title = "Marker Expression in T-cell Clusters", x = "Genes", y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- FeaturePlot(seu_merged, features = "T_Cell_Exhaustion_Score1") +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu"))) +
  ggtitle("Global T-Cell Exhaustion Score")

meta_ds1 <- read.delim("Data/scRNA_GSE140228/GSE140228_UMI_counts_Droplet_cellinfo.tsv", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds1) <- paste0("DS1_", meta_ds1$Barcode)
condition_ds1 <- meta_ds1[, "Tissue", drop = FALSE]
colnames(condition_ds1) <- "Condition"

meta_ds2 <- read.delim("Data/scRNA_GSE149614/GSE149614_HCC.metadata.updated.txt", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds2) <- paste0("DS2_", meta_ds2$Cell)
condition_ds2 <- meta_ds2[, "site", drop = FALSE]
colnames(condition_ds2) <- "Condition"

combined_meta <- rbind(condition_ds1, condition_ds2)
seu_merged <- AddMetaData(seu_merged, metadata = combined_meta)

seu_merged$Condition[seu_merged$Condition == "Adjacent"] <- "Normal"

seu_filtered <- subset(seu_merged, subset = Condition %in% c("Tumor", "Normal"))

p3 <- VlnPlot(seu_filtered, features = "T_Cell_Exhaustion_Score1", 
              idents = target_clusters, group.by = "Condition", pt.size = 0) +
  ggtitle("Exhaustion Score: Normal vs Tumor") +
  stat_compare_means(method = "wilcox.test")

print(p1)
print(p2)
print(p3)
```

### Results: scRNA-Seq

* **Result comparison (prediction #3)**: The scRNA-seq results support our prediction. The DotPlot shows that CD8+ T-cell clusters (specifically 14 and 17) co-express the inhibitory receptors we checked: PDCD1 (PD-1), HAVCR2 (TIM-3), and TIGIT, along with the transcription factor TOX. The VlnPlot confirms that the overall T-Cell Exhaustion Score is significantly higher in the tumor microenvironment than in normal adjacent tissue (Wilcoxon test, $p < 2.2e-16$). This demonstrates that T-cells inside the tumor are in an exhausted state.
* **Biological Integration**: Our findings point to a specific immune evasion mechanism in HCC. Even though CD8+ T-cells manage to infiltrate the tumor, the local environment suppresses their activity. By upregulating inhibitory receptors like PD-1 and TIM-3, the tumor blocks the T-cells' ability to attack cancer cells. Ultimately, this explains how the tumor continues to grow even after an adaptive immune response has been triggered.
