---
title: "Transcriptomic Signatures of Innate Immune Suppression and T-Cell Exhaustion in Hepatocellular Carcinoma"
author: "Re'em Melamed-Katz (213874373), Eliav Katzenstein (326986668)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "Figures/")
```

```{r libraries}
library(clusterProfiler)
library(data.table)
library(DESeq2)
library(EnhancedVolcano)
library(enrichplot)
library(GEOquery)
library(ggplot2)
library(ggpubr)
library(harmony)
library(Matrix)
library(msigdbr)
library(org.Hs.eg.db) # this is the pathway library for a human (for GSEA)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(Seurat)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(tidyverse)
library(xCell2)
```

------------------------------------------------------------------------

## 1. Abstract (300 words in english)

------------------------------------------------------------------------

## 2. Main analysis: Bulk RNA-seq

### 2.1 Prediction #1:

**Prediction**:

**Biological reasoning**:

### 2.2 Decision checkpoint #1:

**The decision**:

**Why we chose it**:

**Alternative considered**:

**Why we rejected the alternative**:

## 2.4 Code:

### Decision checkpoint #1:

This is actually the first decision checkpoint. The choice of TCGA-LIHC
as the dataset is crucial, seeing as it is very general and
comprehensive with many samples (though only with 50 normal samples). An
alternative we had was GSE-144269. But seeing as it was smaller and only
had patients with prior HepD that interfered with the results.

```{r}
# get the data
Bio.LIHC.query <- GDCquery(project = "TCGA-LIHC", 
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      experimental.strategy = "RNA-Seq")
Bio.LIHC.query.res <- getResults(Bio.LIHC.query)
```

```{r}
# download
options(timeout=1000)

GDCdownload(Bio.LIHC.query, directory = "./")
Bio.LIHC.data <- GDCprepare(Bio.LIHC.query, directory = "./")
```

```{r}
# get the LIHC metadata
Bio.LIHC.metadata <- colData(Bio.LIHC.data)
Bio.LIHC.metadata_df <- as.data.frame(Bio.LIHC.metadata)
```

```{r}
# check the data to make sure its clean

# NA values:
NAcolls <- colnames(Bio.LIHC.metadata_df)[sapply(Bio.LIHC.metadata_df, function(x) any(is.na(x)))]
# note that the definition of the sample has no NA so we will not clean anything for now

# amount of normal and cancerous samples:
normalSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "NT")
TumorSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "TP")

# total samples:
SampleCount <- length(Bio.LIHC.metadata_df$shortLetterCode)

# that doesnt add up!
notNormal_or_Tumor <- rownames(Bio.LIHC.metadata_df)[Bio.LIHC.metadata_df$shortLetterCode != "NT" & Bio.LIHC.metadata_df$shortLetterCode != "TP"]

Bio.LIHC.metadata_df[notNormal_or_Tumor, "shortLetterCode"]

# that is a recurrent tumor, which we don't want to deal with so i will remove them
Bio.LIHC_shortLetterCode.clean_metadata_df <- Bio.LIHC.metadata_df[Bio.LIHC.metadata_df$shortLetterCode != "TR",]

# how many NA are in the cancer stage?
sum(is.na(Bio.LIHC.metadata_df$ajcc_pathologic_stage))

# remove NA for stage analysis
Bio.LIHC_Stage.clean_metadata_df <- Bio.LIHC.metadata_df[!is.na(Bio.LIHC.metadata_df$ajcc_pathologic_stage),]

# simplify stages
Bio.LIHC_Stage.clean_metadata_df$simplified_stage <- gsub("[ABC]$", "", Bio.LIHC_Stage.clean_metadata_df$ajcc_pathologic_stage)  

# we want to compare only cancer samples when looking at the cancer stage
Bio.LIHC_Stage.clean_metadata_df <- Bio.LIHC_Stage.clean_metadata_df[Bio.LIHC_Stage.clean_metadata_df$shortLetterCode != "NT", ]
# how many are there from every stage?
sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage 0")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage I")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage II")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage III")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage IV")

# there is only one 0 stage sample. so we will remove it and use only the others
Bio.LIHC_Stage.clean_metadata_df <- Bio.LIHC_Stage.clean_metadata_df[Bio.LIHC_Stage.clean_metadata_df$simplified_stage != "Stage 0",]
```

```{r}
# get the Seq counts data

counts.LIHC <- assay(Bio.LIHC.data)
clean_counts.LIHC_shortLetterCode <- counts.LIHC[,rownames(Bio.LIHC_shortLetterCode.clean_metadata_df)]

clean_counts.LIHC_Stage <- counts.LIHC[,rownames(Bio.LIHC_Stage.clean_metadata_df)]

# check if everything is ready for DESeq
all(rownames(Bio.LIHC_shortLetterCode.clean_metadata_df) == colnames(clean_counts.LIHC_shortLetterCode))

all(rownames(Bio.LIHC_Stage.clean_metadata_df) == colnames(clean_counts.LIHC_Stage))

sum(is.na(Bio.LIHC_shortLetterCode.clean_metadata_df$shortLetterCode))
sum(is.na(Bio.LIHC_Stage.clean_metadata_df$simplified_stage))

```

### Decision checkpoint #2:

We decided to use DESeq2 as our differential expression tool because it
is the one we were most familiar with, a standard and reliable tool.

```{r}
# create DESeq object for the short letter code
LIHC_shortLetterCode.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC_shortLetterCode,
                              colData = Bio.LIHC_shortLetterCode.clean_metadata_df,
                              design = ~ shortLetterCode
)
LIHC_shortLetterCode.dds <- LIHC_shortLetterCode.dds[(rowSums(counts(LIHC_shortLetterCode.dds)) >= 10), ]

# create DESeq object for the stages
LIHC_Stage.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC_Stage,
                              colData = Bio.LIHC_Stage.clean_metadata_df,
                              design = ~ simplified_stage
)
LIHC_Stage.dds <- LIHC_Stage.dds[(rowSums(counts(LIHC_Stage.dds)) >= 10), ]
```

# define reference level

```{r}
# reference level is I, not 0
LIHC_Stage.dds$simplified_stage <- relevel(LIHC_Stage.dds$simplified_stage, ref = "Stage I")

LIHC_shortLetterCode.dds$shortLetterCode <- relevel(LIHC_shortLetterCode.dds$shortLetterCode, ref = "NT")
```

# run DESeq

```{r}
LIHC_shortLetterCode.dds <- DESeq(LIHC_shortLetterCode.dds)
resultsNames(LIHC_shortLetterCode.dds)

LIHC_Stage.dds <- DESeq(LIHC_Stage.dds)
resultsNames(LIHC_Stage.dds)
```

# get resaults

```{r}
res_NT_vs_TP.LFC <- lfcShrink(LIHC_shortLetterCode.dds, coef = "shortLetterCode_TP_vs_NT", type = "apeglm")

res_I_vs_II.LFC <- lfcShrink(LIHC_Stage.dds, coef = "simplified_stage_Stage.II_vs_Stage.I", type = "apeglm")

res_I_vs_III.LFC <- lfcShrink(LIHC_Stage.dds, coef = "simplified_stage_Stage.III_vs_Stage.I", type = "apeglm")

res_I_vs_IV.LFC <- lfcShrink(LIHC_Stage.dds, coef = "simplified_stage_Stage.IV_vs_Stage.I", type = "apeglm")


# clear the genes of their version number (important so GSEA will know what to do and so we can add the gene symbol)
rownames(res_NT_vs_TP.LFC) <- gsub("\\..*", "", 
rownames(res_NT_vs_TP.LFC))

rownames(res_I_vs_II.LFC) <- gsub("\\..*", "", 
rownames(res_I_vs_II.LFC))

rownames(res_I_vs_III.LFC) <- gsub("\\..*", "", rownames(res_I_vs_III.LFC))

rownames(res_I_vs_IV.LFC) <- gsub("\\..*", "", rownames(res_I_vs_IV.LFC))



# add symbols so we can "easily" interpret the genes
# Add gene symbols
res_NT_vs_TP.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_NT_vs_TP.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

res_I_vs_II.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_I_vs_II.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

res_I_vs_III.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_I_vs_III.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")
res_I_vs_IV.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_I_vs_IV.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

```

# perform pca to visualize a bit whats going on

```{r}
any(duplicated(rownames(res_NT_vs_TP.LFC)))
any(duplicated(rownames(res_I_vs_II.LFC)))
any(duplicated(rownames(res_I_vs_III.LFC)))
any(duplicated(rownames(res_I_vs_IV.LFC)))

counts.vst.LIHC_shortLetterCode <- DESeq2::vst(LIHC_shortLetterCode.dds, blind = TRUE)

pcaData.LIHC_shortLetterCode <- DESeq2::plotPCA(counts.vst.LIHC_shortLetterCode, ntop = 1000, intgroup = c("shortLetterCode"), returnData = TRUE)

counts.vst.LIHC_Stage <- DESeq2::vst(LIHC_Stage.dds, blind = TRUE)

pcaData.LIHC_Stage <- DESeq2::plotPCA(counts.vst.LIHC_Stage, ntop = 1000, intgroup = c("simplified_stage"), returnData = TRUE)
```
### prediction #1:
It is safe to say that we will see a lot of variance between the tumor and non tumor samples, seeing as the biological difference between them is sky and land.
The variance between the stages should be seen, but probably not as much compared to the tumor and non tumor samples.
# visualize

```{r}
percentVar <- round(100 * attr(pcaData.LIHC_shortLetterCode, "percentVar"))
plot_shortLetterCode <- ggplot(pcaData.LIHC_shortLetterCode, aes(x=PC1, y=PC2, color=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))

percentVar <- round(100 * attr (pcaData.LIHC_Stage, "percentVar"))
plot_Stage <- ggplot(pcaData.LIHC_Stage, aes(x=PC1, y=PC2, color=simplified_stage)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))

combined_plot <- ggplot(pcaData.LIHC_Stage, aes(x=PC1, y=PC2, color=simplified_stage, shape=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))
combined_plot
plot_shortLetterCode
plot_Stage
# looks awesome!
```
These plots look the same (except for the flip) for an important reason. and that is that PCA does not care for our design, and just looks at the two biggest directions of varience.
We can see a cluster of non tumor samples up to the bottom right. But weirdly enough, all the different stages are mixed together. This could suggest that the principal componants of the variance between nonTumor and Tumor samples is not at all related to the differences between the stages.

# look at the volcano plot

```{r fig.width=8, fig.height=8}

EnhancedVolcano::EnhancedVolcano(res_I_vs_II.LFC,
                lab = res_I_vs_II.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.05,
                title = 'Stage I vs Stage II cancer')
EnhancedVolcano::EnhancedVolcano(res_I_vs_IV.LFC,
                lab = res_I_vs_IV.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.05,
                title = 'Stage I vs Stage IV cancer')

EnhancedVolcano::EnhancedVolcano(res_NT_vs_TP.LFC,
                lab = res_NT_vs_TP.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.05,
                title = 'Non Tumor vs Tumor samples')


```

lets look at some genes with high log_2 fold change and p value from the NT to TP volcano plot:
TERT - an important gene that helps in the production of telomerase which when
activated, allows cell reproduction to continue (this is a highly
expressed gene in most cancer, so a good sign)

CPLX2 - this is actually a cognitive related gene, that low expression
of points to neorological decline (high expressivity seems not
connected, but doesnt necciseraly mean anything)

LINC01419 - from what i read this gene literally just ruins everything
and helps the cancer. in the nucleus it supresses tumor supressing
genes. and it ruins somehow the DNA repair, which also helps tumor grow
(one could say that the cancer is here because of the high expression,
but im not going to take that jump)



### Desicion checkpoint #3:

Seeing as we want to check the immune pathways (as in, know a lot about
a little), GSEA is an excellent tool for the job. By choosing specific
pathways (as seen later), we can see excatly how and how much is that
pathway differentially expressed. An alternative was ORA, but it is much
less sophisticated and robust, and it is much more suited for quick
small answers, while we need this as a robust and main part of our
analysis.

### GSEA

```{r}
# sort the genes by differentiallity


# get differentiallity
oredered_genes.Tumor <- res_NT_vs_TP.LFC$log2FoldChange
oredered_genes.Stage_II <- res_I_vs_II.LFC$log2FoldChange

# check for NA
sum(is.na(oredered_genes.Tumor))
sum(is.na(oredered_genes.Stage_II))

# give names (cause thats just what you gotta do)
names(oredered_genes.Tumor) <- rownames(res_NT_vs_TP.LFC)
names(oredered_genes.Stage_II) <- rownames(res_I_vs_II.LFC)

# order
oredered_genes.Tumor <- sort(oredered_genes.Tumor, decreasing = TRUE)
oredered_genes.Stage_II <- sort(oredered_genes.Stage_II, decreasing = TRUE)

```
```{r}
# we got some weird artifacts. so we will try a different metric to order the genes with

#res_I_vs_II.LFC$gsea_metric <- sign(res_I_vs_II.LFC$log2FoldChange) * -log10(res_I_vs_II.LFC$pvalue)

#res_clean_I_vs_II <- res_I_vs_II.LFC[!is.na(res_I_vs_II.LFC$gsea_metric),]

# order by this new metric
#oredered_genes.Stage_II <- res_clean_I_vs_II$gsea_metric
#names(oredered_genes.Stage_II) <- rownames(res_clean_I_vs_II)
#oredered_genes.Stage_II <- sort(oredered_genes.Stage_II, decreasing = TRUE)
```

### Decision checkpoint #4:

A p-value of 0.05 was chosen as a standard p-value for statistical
significance

```{r}
# perform GSEA (using hallmark sets for a human)
hallmark <- msigdbr(species = "Homo sapiens") %>% dplyr::select(gs_name, ensembl_gene)

gseaHallmark_res.Tumor <- GSEA(geneList = oredered_genes.Tumor,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)
```
```{r}
gseaHallmark_res.Stage_II <- GSEA(geneList = oredered_genes.Stage_II,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)

```

```{r}
# visualize

dotplot(gseaHallmark_res.Stage_II, showCategory = 10, split = ".sign", 
        title = "tumor sample enriched patways") +
   theme_minimal(base_size = 6) + facet_grid(. ~ .sign)
```

so there are really enriched and supressed genes (mostly supressed it
seems). but we want to specifically look at the effect on immune
pathways



### Prediction #1:
A prediction of the enrichment score of these 4 immune pathways can go both ways.
We would assume that there will be variance, meaning the immune pathways are either supressed or activated.
As learned during the tutorials, this will point towards the tumor being "hot" or "cold" meaning an active immune environment and a supressed immune environment respectively.
```{r}
enrichplot_Inflammatory_response <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response")

enrichplot_Interferon_alpha_response <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response")

enrichplot_Interferon_gamma_response <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response")

enrichplot_TNFA_signaling <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling")


enrichplot_combined = enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = c("HALLMARK_INFLAMMATORY_RESPONSE",
                                    "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                    "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                                    "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
                                    ),
                      title = "TNFA signaling")
enrichplot_Inflammatory_response
enrichplot_Interferon_alpha_response
enrichplot_Interferon_gamma_response
enrichplot_TNFA_signaling
#enrichplot_combined
```

all four pathways are an integral part of the immune system. When
looking at the enrichment score graph, we can see that the genes are
differentially suppressed in the tumor microenvironment. In this case,
we say that the tumor is "cold" meaning, it suppresses immune activity.

### xCell analysis

so we know that the immune system is suppressed around this tumor. but
we want to know exactly what cells are suppressed. for that we use
xCell2 and its immune reference.

```{r}
# first we get the normalized  counts from the differential expression
normalized_counts_matrix.LIHC_shortLetterCode <- counts(LIHC_shortLetterCode.dds, normalized = TRUE)
normalized_counts_matrix.LIHC_Stage <- counts(LIHC_Stage.dds, normalized = TRUE)

sum(is.na(rownames(normalized_counts_matrix.LIHC_shortLetterCode)))
sum(is.na(rownames(normalized_counts_matrix.LIHC_Stage)))

# remove version number
rownames(normalized_counts_matrix.LIHC_shortLetterCode) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC_shortLetterCode))

rownames(normalized_counts_matrix.LIHC_Stage) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC_Stage))

# make symbol (important to do twice because they dont have excatly the same data)
gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC_shortLetterCode),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC_shortLetterCode) <- gene_symbols[rownames(normalized_counts_matrix.LIHC_shortLetterCode)]

gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC_Stage),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC_Stage) <- gene_symbols[rownames(normalized_counts_matrix.LIHC_Stage)]

# no error! that's a relief. lets check for NA
sum(is.na(rownames(normalized_counts_matrix.LIHC_shortLetterCode)))
sum(is.na(rownames(normalized_counts_matrix.LIHC_Stage)))

# thats a lot of na, so lets get them out
clean_normalized_counts_matrix.LIHC_shortLetterCode <- normalized_counts_matrix.LIHC_shortLetterCode[!is.na(rownames(normalized_counts_matrix.LIHC_shortLetterCode)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC_shortLetterCode)))

clean_normalized_counts_matrix.LIHC_Stage <- normalized_counts_matrix.LIHC_Stage[!is.na(rownames(normalized_counts_matrix.LIHC_Stage)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC_Stage)))
```



```{r}
# start the actual analysis

# load immune reference
data(ImmuneCompendium.xCell2Ref, package = "xCell2")

# run analysis
xCell_immune_res.LIHC_shortLetterCode <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC_shortLetterCode,
                               xcell2object = ImmuneCompendium.xCell2Ref)

xCell_immune_res.LIHC_Stage <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC_Stage,
                               xcell2object = ImmuneCompendium.xCell2Ref)
```


### Prediction #2:
Seeing as the 4 immune pathways we chose are suppressed around the tumor, we would assume that all if not some of the immune cells connected to that pathway would be supressed.

Inflammatory Response: "Macrophages", "Neutrophils", "T-Cells", "B-Cells", "NK-Cells", "D-Cells"
Interfron Alpha Response: "NK-Cells", "B-Cells", "T-Cells", "D-Cells"
Interfron Gamma Response: "Macrophages", "T Helper Cells", "CD8+ Cytotoxic T-Cells", "NK-Cells", "B-Cells", "D-Cells"
TNFA signaling: "Macrophages", "T-Cells", "B-Cells", "NK-Cells", "Fibroblasts", "Keratinocytes"
```{r fig.width=8, fig.height=8}
# lets visualize the results

# we need an "annotation col" for some reason
annotation_col <- data.frame(Condition = Bio.LIHC.clean_metadata_df$shortLetterCode)
rownames(annotation_col) <- rownames(Bio.LIHC.clean_metadata_df)

# order the samples by sample type
ordered_for_heatmap <- rownames(Bio.LIHC.clean_metadata_df)[order(Bio.LIHC.clean_metadata_df$shortLetterCode)]
ordered_xCell2_res <- xCell_immune_res.LIHC[, ordered_for_heatmap]
pheatmap(ordered_xCell2_res, 
         annotation_col = annotation_col,
         scale = "row",
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         breaks = seq(-2, 2, length.out = 100),
         cutree_rows = 5,
         main = "xCell2 cell type heatmap")
```

we see a clear immune deactivation in the tumor samples compared to the
non tumor samples, when looking at: - granulocyte - natural killer cell -
monocyte - neutrophil - plasmacytoid dendritic cell - non classical
monocyte - eosinophil - classical nonocyte - basophil and more with
names too long. 
by looking at the hierarcial clustering and cutoff, we
can see that all supressed immune cells are closely related to each
other. and by asking gemini (don't know what else i can use), these are
all cells that are a big part of the "innate immune system" who's job is
to be the "first line of defense" when there is some type of attack.
When the Innate Immune System is activated for too long, the Adaptive
Immune System comes in. But with the Innate Immune System being active
in normal samples and not tumor ones could point maybe towards an immune
response, but not where the tumor is.

### 2.5 Results:

**Result comparison (prediction #1)**:

**Biological Integration**:

------------------------------------------------------------------------

## 3. Secondary analysis: scRNA-Seq

### 3.1 Prediction #2

**Prediction**:

**Biological reasoning**:

### 2.3 Decision checkpoint #2:

**The decision**:

**Why we chose it**:

**Alternative considered**:

**Why we rejected the alternative**:

### 3.2 Code:

```{r harmony}
# load 1st dataset
sc_data_1 <- ReadMtx(
  mtx = "Data/scRNA_GSE140228/matrix.mtx",
  cells = "Data/scRNA_GSE140228/barcodes.tsv",
  features = "Data/scRNA_GSE140228/features.tsv",
  skip.feature = 1
)
seu_1 <- CreateSeuratObject(counts = sc_data_1, project = "GSE140228", min.cells = 3, min.features = 200)

# load 2nd dataset
file_path_2 <- "Data/scRNA_GSE149614/GSE149614_HCC.scRNAseq.S71915.count.txt"
sc_data_2_raw <- fread(file_path_2, data.table = FALSE, select = 1:3001)
rownames(sc_data_2_raw) <- sc_data_2_raw[, 1]
sc_data_2_clean <- sc_data_2_raw[, -1]
sc_data_2_sparse <- as(as.matrix(sc_data_2_clean), "dgCMatrix")

seu_2 <- CreateSeuratObject(counts = sc_data_2_sparse, project = "GSE149614", min.cells = 3, min.features = 200)

# merge and filter out low quality cells
seu_merged <- merge(seu_1, y = seu_2, add.cell.ids = c("DS1", "DS2"), project = "HCC_Integrated")
seu_merged[["percent.mt"]] <- PercentageFeatureSet(seu_merged, pattern = "^MT-")
seu_merged <- subset(seu_merged, subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mt < 10)

# pipeline and Harmony batch correction
seu_merged <- NormalizeData(seu_merged) %>% 
              FindVariableFeatures() %>% 
              ScaleData() %>% 
              RunPCA() %>% 
              RunHarmony(group.by.vars = "orig.ident")

# clustering and UMAP
seu_merged <- FindNeighbors(seu_merged, reduction = "harmony", dims = 1:15)
seu_merged <- FindClusters(seu_merged, resolution = 0.5)
seu_merged <- RunUMAP(seu_merged, reduction = "harmony", dims = 1:15)

# visual check plots
DimPlot(seu_merged, reduction = "umap", label = TRUE) + ggtitle("Cluster Map")
FeaturePlot(seu_merged, features = "CD8A") + ggtitle("CD8+ T-cell Localization")
```

We examined the UMAP and CD8A FeaturePlot to localize the T-cell
compartment, and identified clusters 0, 1, 6, 14, 17 as our target
populations.

### 3.2 Decision checkpoint #3: cluster characterization method

**The decision**: We chose to use logistic regression with dataset
integration.

**Why we chose it**: Although we integrated our datasets using Harmony,
Harmony only corrects the low dimensional embedding (PCA and UMAP) to
allow for shared clustering. It does not modify the underlying gene
expression matrix. By using logistic regression with latent.vars =
"orig.ident", we can explicitly account for and neutralize the batch
effects (technical variation between the two studies) during the actual
marker discovery. This ensures our markers are biologically driven by
cell type and not artifacts of different laboratories.

**Alternative considered**: Wilcoxon rank-sum test.

**Why we rejected the alternative**: The Wilcoxon test is a
non-parametric test that compares ranks but cannot incorporate
covariates. Using it on merged datasets without accounting for batch
effects often leads to "false positive" markers that represent dataset
differences rather than true cell-type biology.

### 3.3 Code:

```{r lr}
target_clusters <- c(0, 1, 6, 14, 17)

# join all layers
seu_merged <- JoinLayers(seu_merged)

# extract our T-cell group markers
t_markers <- FindMarkers(seu_merged, 
                         ident.1 = target_clusters, 
                         test.use = "LR", 
                         latent.vars = "orig.ident", 
                         only.pos = TRUE, 
                         min.pct = 0.25)

# keep only the significant genes
sig_genes <- rownames(t_markers[t_markers$p_val_adj < 0.05 & t_markers$avg_log2FC > 0.5, ])

# GO
ego_results <- enrichGO(gene          = sig_genes,
                        OrgDb         = org.Hs.eg.db,
                        keyType       = 'SYMBOL',
                        ont           = "BP", 
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

dotplot(ego_results, showCategory = 15, label_format = 40) + 
  theme_bw() + 
  scale_color_gradient(low = "red", high = "blue") +
  ggtitle("Functional Enrichment: CD8+ T-cell Compartment") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

Now, we will examine if our target T-cells actively co-express classical
exhaustion markers. We will also test if this overall exhaustion
signature is significantly stronger in the tumor environment compared to
normal tissue:

```{r results}
exhaustion_list <- list(c("PDCD1", "LAG3", "HAVCR2", "TIGIT", "CTLA4", "TOX", "LAYN"))

seu_merged <- AddModuleScore(
  object = seu_merged,
  features = exhaustion_list,
  name = "T_Cell_Exhaustion_Score"
)

exhaustion_markers <- c("CD8A", "PDCD1", "LAG3", "HAVCR2", "TIGIT", "TOX")

p1 <- DotPlot(seu_merged, features = exhaustion_markers, idents = target_clusters) +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu"))) +
  labs(title = "Marker Expression in T-cell Clusters", x = "Genes", y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- FeaturePlot(seu_merged, features = "T_Cell_Exhaustion_Score1") +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu"))) +
  ggtitle("Global T-Cell Exhaustion Score")

meta_ds1 <- read.delim("Data/scRNA_GSE140228/GSE140228_UMI_counts_Droplet_cellinfo.tsv", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds1) <- paste0("DS1_", meta_ds1$Barcode)
condition_ds1 <- meta_ds1[, "Tissue", drop = FALSE]
colnames(condition_ds1) <- "Condition"

meta_ds2 <- read.delim("Data/scRNA_GSE149614/GSE149614_HCC.metadata.updated.txt", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds2) <- paste0("DS2_", meta_ds2$Cell)
condition_ds2 <- meta_ds2[, "site", drop = FALSE]
colnames(condition_ds2) <- "Condition"

combined_meta <- rbind(condition_ds1, condition_ds2)
seu_merged <- AddMetaData(seu_merged, metadata = combined_meta)

seu_merged$Condition[seu_merged$Condition == "Adjacent"] <- "Normal"

seu_filtered <- subset(seu_merged, subset = Condition %in% c("Tumor", "Normal"))

p3 <- VlnPlot(seu_filtered, features = "T_Cell_Exhaustion_Score1", 
              idents = target_clusters, group.by = "Condition", pt.size = 0) +
  ggtitle("Exhaustion Score: Normal vs Tumor") +
  stat_compare_means(method = "wilcox.test")

print(p1)
print(p2)
print(p3)
```

### 3.4 Results:

**Result comparison (prediction #2)**:

**Biological Integration**:
