---
title: "Transcriptomic Signatures of Innate Immune Suppression and T-Cell Exhaustion in Hepatocellular Carcinoma"
author: "Re'em Melamed-Katz (213874373), Eliav Katzenstein (326986668)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "Figures/")
```

```{r libraries}
library(clusterProfiler)
library(data.table)
library(DESeq2)
library(EnhancedVolcano)
library(enrichplot)
library(GEOquery)
library(ggplot2)
library(ggpubr)
library(harmony)
library(Matrix)
library(msigdbr)
library(org.Hs.eg.db) # this is the pathway library for a human (for GSEA)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(Seurat)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(tidyverse)
library(xCell2)
```


## 1. Abstract (300 words in english)

Hepatocellular carcinoma (HCC) is the most common liver cancer, and one of the most common cancers overall. It is the third leading cause of cancer related deaths, and is especially common among those with liver disease such as cirrhosis and fibrosis.
A better understanding of the immune system and its mechanisms in the tumor microenvironment are needed for better research and understanding of this cancer. Immunotherapy is an upcoming treatment that could also solve the differences between the HCC stages, so knowing the immune system is integral for treating it.
We will look at the differential gene expression between stages of the cancer, using normal tissue samples to give us a general view. Differential expression, Pathway and xCell analysis are used to see how the cancer and its stages affect the immune system. Our secondary analysis Using Seurat takes a closer look at CD8+ T cells, which are a vital part of immunotherapy, and general immune related genes.
Our biggest findings are the inaffect the cancer stage has on the immune microenvironment, the supression of the innate immune system by HCC, and the exhaustion of CD8+ T Cells and PDCD1 (PD-1 encoder).
These findings shine a new light on the current immunotherapy situation, and where energy and research should be directed for better and improved treatments that can be found.

## 2. Main analysis: Bulk RNA-seq

### Data Acqusition and Preprocessing

-   **Decision checkpoint #1: Dataset Selection**
-   **The decision:**: TCGA-LIHC as the dataset.
-   **Why we chose it**: It is crucial, seeing as it is very general and
    comprehensive with many samples (though only with 50 normal
    samples.)
-   **Alternative considered**: GSE-144269
-   **Why we rejected the alternative**: Seeing as it was smaller and
    only had patients with prior HepD that interfered with the results.

```{r}
# get the data
Bio.LIHC.query <- GDCquery(project = "TCGA-LIHC", 
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      experimental.strategy = "RNA-Seq")
Bio.LIHC.query.res <- getResults(Bio.LIHC.query)
```

```{r}
# download
options(timeout=1000)

GDCdownload(Bio.LIHC.query, directory = "./")
Bio.LIHC.data <- GDCprepare(Bio.LIHC.query, directory = "./")
```

```{r}
# get the LIHC metadata
Bio.LIHC.metadata <- colData(Bio.LIHC.data)
Bio.LIHC.metadata_df <- as.data.frame(Bio.LIHC.metadata)
```

```{r}
# check the data to make sure its clean

# NA values:
NAcolls <- colnames(Bio.LIHC.metadata_df)[sapply(Bio.LIHC.metadata_df, function(x) any(is.na(x)))]
# note that the definition of the sample has no NA so we will not clean anything for now

# amount of normal and cancerous samples:
normalSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "NT")
TumorSampleCount <- sum(Bio.LIHC.metadata_df$shortLetterCode == "TP")

# total samples:
SampleCount <- length(Bio.LIHC.metadata_df$shortLetterCode)

# that doesnt add up!
notNormal_or_Tumor <- rownames(Bio.LIHC.metadata_df)[Bio.LIHC.metadata_df$shortLetterCode != "NT" & Bio.LIHC.metadata_df$shortLetterCode != "TP"]

Bio.LIHC.metadata_df[notNormal_or_Tumor, "shortLetterCode"]

# that is a recurrent tumor, which we don't want to deal with so i will remove them
Bio.LIHC_shortLetterCode.clean_metadata_df <- Bio.LIHC.metadata_df[Bio.LIHC.metadata_df$shortLetterCode != "TR",]

# how many NA are in the cancer stage?
sum(is.na(Bio.LIHC.metadata_df$ajcc_pathologic_stage))

# remove NA for stage analysis
Bio.LIHC_Stage.clean_metadata_df <- Bio.LIHC.metadata_df[!is.na(Bio.LIHC.metadata_df$ajcc_pathologic_stage),]

# simplify stages
Bio.LIHC_Stage.clean_metadata_df$simplified_stage <- gsub("[ABC]$", "", Bio.LIHC_Stage.clean_metadata_df$ajcc_pathologic_stage)  

# we want to compare only cancer samples when looking at the cancer stage
Bio.LIHC_Stage.clean_metadata_df <- Bio.LIHC_Stage.clean_metadata_df[Bio.LIHC_Stage.clean_metadata_df$shortLetterCode != "NT", ]
# how many are there from every stage?
sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage 0")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage I")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage II")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage III")

sum(Bio.LIHC_Stage.clean_metadata_df$simplified_stage == "Stage IV")

# there is only one 0 stage sample. so we will remove it and use only the others
Bio.LIHC_Stage.clean_metadata_df <- Bio.LIHC_Stage.clean_metadata_df[Bio.LIHC_Stage.clean_metadata_df$simplified_stage != "Stage 0",]
```

```{r}
# get the Seq counts data

counts.LIHC <- assay(Bio.LIHC.data)
clean_counts.LIHC_shortLetterCode <- counts.LIHC[,rownames(Bio.LIHC_shortLetterCode.clean_metadata_df)]

clean_counts.LIHC_Stage <- counts.LIHC[,rownames(Bio.LIHC_Stage.clean_metadata_df)]

# check if everything is ready for DESeq
all(rownames(Bio.LIHC_shortLetterCode.clean_metadata_df) == colnames(clean_counts.LIHC_shortLetterCode))

all(rownames(Bio.LIHC_Stage.clean_metadata_df) == colnames(clean_counts.LIHC_Stage))

sum(is.na(Bio.LIHC_shortLetterCode.clean_metadata_df$shortLetterCode))
sum(is.na(Bio.LIHC_Stage.clean_metadata_df$simplified_stage))

```

### Differential Expression Analysis

-   **Decision checkpoint #2: Differential Expression Tool**
-   **The decision**: We decided to use DESeq2 as our differential
    expression tool.
-   **Why we chose it**: It is the one we were most familiar with, a
    standard and reliable tool.
-   **Alternative considered**: edgeR.
-   **Why we rejected the alternative**: Preferred DESeq2 for
    familiarity.

```{r}
# create DESeq object for the short letter code
LIHC_shortLetterCode.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC_shortLetterCode,
                              colData = Bio.LIHC_shortLetterCode.clean_metadata_df,
                              design = ~ shortLetterCode
)
LIHC_shortLetterCode.dds <- LIHC_shortLetterCode.dds[(rowSums(counts(LIHC_shortLetterCode.dds)) >= 10), ]

# create DESeq object for the stages
LIHC_Stage.dds <- DESeqDataSetFromMatrix(
                              countData = clean_counts.LIHC_Stage,
                              colData = Bio.LIHC_Stage.clean_metadata_df,
                              design = ~ simplified_stage
)
LIHC_Stage.dds <- LIHC_Stage.dds[(rowSums(counts(LIHC_Stage.dds)) >= 10), ]
```

**define reference level**

```{r}
# reference level is I, not 0
LIHC_Stage.dds$simplified_stage <- relevel(LIHC_Stage.dds$simplified_stage, ref = "Stage I")

LIHC_shortLetterCode.dds$shortLetterCode <- relevel(LIHC_shortLetterCode.dds$shortLetterCode, ref = "NT")
```

**run DESeq**

```{r}
LIHC_shortLetterCode.dds <- DESeq(LIHC_shortLetterCode.dds)
resultsNames(LIHC_shortLetterCode.dds)

LIHC_Stage.dds <- DESeq(LIHC_Stage.dds)
resultsNames(LIHC_Stage.dds)
```

**get results**

```{r}
res_NT_vs_TP.LFC <- lfcShrink(LIHC_shortLetterCode.dds, coef = "shortLetterCode_TP_vs_NT", type = "apeglm")

res_I_vs_II.LFC <- lfcShrink(LIHC_Stage.dds, coef = "simplified_stage_Stage.II_vs_Stage.I", type = "apeglm")

res_I_vs_III.LFC <- lfcShrink(LIHC_Stage.dds, coef = "simplified_stage_Stage.III_vs_Stage.I", type = "apeglm")

res_I_vs_IV.LFC <- lfcShrink(LIHC_Stage.dds, coef = "simplified_stage_Stage.IV_vs_Stage.I", type = "apeglm")


# clear the genes of their version number (important so GSEA will know what to do and so we can add the gene symbol)
rownames(res_NT_vs_TP.LFC) <- gsub("\\..*", "", 
rownames(res_NT_vs_TP.LFC))

rownames(res_I_vs_II.LFC) <- gsub("\\..*", "", 
rownames(res_I_vs_II.LFC))

rownames(res_I_vs_III.LFC) <- gsub("\\..*", "", rownames(res_I_vs_III.LFC))

rownames(res_I_vs_IV.LFC) <- gsub("\\..*", "", rownames(res_I_vs_IV.LFC))



# add symbols so we can "easily" interpret the genes
# Add gene symbols
res_NT_vs_TP.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_NT_vs_TP.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

res_I_vs_II.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_I_vs_II.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

res_I_vs_III.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_I_vs_III.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")
res_I_vs_IV.LFC$symbol <- mapIds(org.Hs.eg.db,
                        keys = rownames(res_I_vs_IV.LFC),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

```

**perform pca to visualize a bit whats going on**

```{r}
# get vst (variance stabilizing transformation)
counts.vst.LIHC_shortLetterCode <- DESeq2::vst(LIHC_shortLetterCode.dds, blind = TRUE)

# apply PCA
pcaData.LIHC_shortLetterCode <- DESeq2::plotPCA(counts.vst.LIHC_shortLetterCode, ntop = 1000, intgroup = c("shortLetterCode"), returnData = TRUE)

counts.vst.LIHC_Stage <- DESeq2::vst(LIHC_Stage.dds, blind = TRUE)

pcaData.LIHC_Stage <- DESeq2::plotPCA(counts.vst.LIHC_Stage, ntop = 1000, intgroup = c("simplified_stage"), returnData = TRUE)
```

**Prediction #1: PCA dotplot clusters**

-   **Prediction**: All the non tumor samples will cluster and all the
    tumor samples will cluster. The stages of the cancer will also
    cluster together, but more loosely.
-   **Biological Reasoning**: The difference between a tumor sample and
    non tumor sample is *huge*, a cancer is like another organism
    growing inside you. The differences between stages of the cancer
    should exist but be mild compared to the differences between tumor
    and non tumor tissue.

```{r}
percentVar <- round(100 * attr(pcaData.LIHC_shortLetterCode, "percentVar"))
plot_shortLetterCode <- ggplot(pcaData.LIHC_shortLetterCode, aes(x=PC1, y=PC2, color=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))

percentVar <- round(100 * attr (pcaData.LIHC_Stage, "percentVar"))
plot_Stage <- ggplot(pcaData.LIHC_Stage, aes(x=PC1, y=PC2, color=simplified_stage)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))

combined_plot <- ggplot(pcaData.LIHC_Stage, aes(x=PC1, y=PC2, color=simplified_stage, shape=shortLetterCode)) + geom_point(size = 3) + labs(
       subtitle = "Each point is a patient's gene expression profile",
       x = paste0("PC1 (", percentVar[1], "% Variance)"),
       y = paste0("PC2 (", percentVar[2], "% Variance)"))
plot_shortLetterCode
plot_Stage
# looks awesome!
```

-   **Result comparison (prediction #1)**:That is interesting! As predicted, non
    tumor samples all clustered together, but contrary to what we
    thought, the tumor samples were all over the place (even when
    looking through the stages). After some searching, it seems that the
    reason is quite simple. Cancer is a very varied disease, the
    differnces between the same cancer in two different people can be
    very big! And so it makes sense that the tumor samples will "spread
    out". The fact that the Stage wasn't a big principal componant is a
    bit weird, but also makes sense when thinking of the fact that
    different people with different stages of cancer can vary a lot, and
    are very dependant on the patient.

**look at the volcano plot**

**note about the plots**: The p-value cutoff for all plots are 0.05 as a
standard p-value for statistical significance. But the FC cutoff is
different. Because we expect massive differences between tumor and non
tumor samples, the fold change is naturally big for many of the genes,
so the FC cutoff for the Tumor_vs_Non_Tumor plot is 5. While for the
stage differnces, it is 2.5 .

Also note that the volcano plot of Stage
I to Stage IV looks weird, this is probably due to the fact that only 3
stage IV samples are present, which may cause some anomalies.

```{r fig.width=8, fig.height=8}

EnhancedVolcano::EnhancedVolcano(res_I_vs_II.LFC,
                lab = res_I_vs_II.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 2.5, 
                pCutoff = 0.05,
                title = 'Stage I vs Stage II cancer')
EnhancedVolcano::EnhancedVolcano(res_I_vs_III.LFC,
                lab = res_I_vs_III.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 2.5, 
                pCutoff = 0.05,
                title = 'Stage I vs Stage III cancer')
EnhancedVolcano::EnhancedVolcano(res_I_vs_IV.LFC,
                lab = res_I_vs_IV.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 2.5, 
                pCutoff = 0.05,
                title = 'Stage I vs Stage IV cancer')

EnhancedVolcano::EnhancedVolcano(res_NT_vs_TP.LFC,
                lab = res_NT_vs_TP.LFC$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                labSize = 3,
                FCcutoff = 5, 
                pCutoff = 0.05,
                title = 'Non Tumor vs Tumor samples')


```

lets look at some genes with high log_2 fold change and p value from the
NT to TP volcano plot:

TERT - an important gene that helps in the
production of telomerase which when activated, allows cell reproduction
to continue (this is a highly expressed gene in most cancer, so a good
sign) 

CPLX2 - this is actually a cognitive related gene, that low
expression of points to neorological decline (high expressivity seems
not connected, but doesnt necciseraly mean anything) 

LINC01419 - from
what i read this gene literally just ruins everything and helps the
cancer. in the nucleus it supresses tumor supressing genes. and it ruins
somehow the DNA repair, which also helps tumor grow (one could say that
the cancer is here because of the high expression, but im not going to
take that jump)

### Pathway Enrichment Analysis

**Desicion checkpoint #3: Pathway Analysis Tool**

-   **The decision**: GSEA is an excellent tool for the job.
-   **Why we chose it**: Seeing as we want to check the immune pathways
    (as in, know a lot about a little), GSEA was built for this exact
    purpous, and is robust and versetile.
-   **Alternative considered**: ORA.
-   **Why we rejected the alternative**: Not only it is less
    sophisticated, and is more well suited for quick small answers, the
    experience and familiarity with GSEA is non replacable.

```{r}
# sort the genes by differentiallity

# note that res_I_vs_IV is ommited here, due to the lack of statistical power from only 3 samples

# get differentiallity
oredered_genes.Tumor <- res_NT_vs_TP.LFC$log2FoldChange
oredered_genes.Stage_II <- res_I_vs_II.LFC$log2FoldChange
oredered_genes.Stage_III <- res_I_vs_III.LFC$log2FoldChange

# check for NA
sum(is.na(oredered_genes.Tumor))
sum(is.na(oredered_genes.Stage_II))
sum(is.na(oredered_genes.Stage_III))

# give names (cause thats just what you gotta do)
names(oredered_genes.Tumor) <- rownames(res_NT_vs_TP.LFC)
names(oredered_genes.Stage_II) <- rownames(res_I_vs_II.LFC)
names(oredered_genes.Stage_III) <- rownames(res_I_vs_III.LFC)

# order
oredered_genes.Tumor <- sort(oredered_genes.Tumor, decreasing = TRUE)
oredered_genes.Stage_II <- sort(oredered_genes.Stage_II, decreasing = TRUE)
oredered_genes.Stage_III <- sort(oredered_genes.Stage_III, decreasing = TRUE)

```

A p-value of 0.05 was chosen as a standard p-value for statistical
significance

```{r}
hallmark <- msigdbr(species = "Homo sapiens") %>% dplyr::select(gs_name, ensembl_gene)
```

```{r}
# perform GSEA (using hallmark sets for a human)


gseaHallmark_res.Tumor <- GSEA(geneList = oredered_genes.Tumor,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)
```

```{r}
gseaHallmark_res.Stage_II <- GSEA(geneList = oredered_genes.Stage_II,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)

```

```{r}
gseaHallmark_res.Stage_III <- GSEA(geneList = oredered_genes.Stage_III,
                                      pvalueCutoff = 0.05,
                                      nPermSimple = 10000,
                                      TERM2GENE = hallmark,
                                      verbose = FALSE, 
                                      eps = 0)

```

**Significant pathways**

**note about the plots**: When running the GSEA
analysis, a p-value cutoff of 0.05 was chosen for all of them for
statistical significance. The plot of the significant pathways in
stage_I_vs_stage_II was ommited seeing as GSEA has found no significant
pathways. This is not concerning or suprising, the fact we got
significant pathways between stage I and stage III is quite suprising.

```{r}
# visualize
dotplot(gseaHallmark_res.Tumor, showCategory = 10, split = ".sign", 
        title = "tumor sample enriched patways") +
   theme_minimal(base_size = 10) + facet_grid(. ~ .sign)

#dotplot(gseaHallmark_res.Stage_II, showCategory = 10, split = ".sign", 
#        title = "tumor sample enriched patways") +
#   theme_minimal(base_size = 10) + facet_grid(. ~ .sign)

dotplot(gseaHallmark_res.Stage_III, showCategory = 10, split = ".sign", 
        title = "stage I_vs_III sample enriched patways") +
   theme_minimal(base_size = 10) + facet_grid(. ~ .sign)

```

so there are really enriched and supressed genes (mostly supressed it
seems). but we want to specifically look at the effect on immune
pathways

**Prediction #2: Immune Pathway Activity**

-   **Prediction**: When talking about the difference between the tumor
    and non tumor samples, the enrichemnt score can go both ways. It is
    safe to assume that the 4 pathways will either be highly activated
    or supressed. And when looking at the different stages, we don't
    think that much supression or activation will occur.
-   **Biological reasoning**: As stated before, the difference between
    tumor and non tumor tissue is massive, so we would expect a massive
    supression or activation of the immune system ("cold" and "hot"
    tumors respectively). And when it comes to the cancer stages, we
    have seen many times that they do not change much themselves.

```{r}
enrichplot_Inflammatory_response_Tumor <- enrichplot::gseaplot2(gseaHallmark_res.Tumor, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response Tumor") 

enrichplot_Interferon_alpha_response_Tumor <- enrichplot::gseaplot2(gseaHallmark_res.Tumor, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response Tumor")

enrichplot_Interferon_gamma_response_Tumor <- enrichplot::gseaplot2(gseaHallmark_res.Tumor, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response Tumor")

enrichplot_TNFA_signaling_Tumor <- enrichplot::gseaplot2(gseaHallmark_res.Tumor, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling Tumor")


enrichplot_Inflammatory_response_II <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response Stage II")

enrichplot_Interferon_alpha_response_II <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response Stage II")

enrichplot_Interferon_gamma_response_II <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response Stage II")

enrichplot_TNFA_signaling_II <- enrichplot::gseaplot2(gseaHallmark_res.Stage_II, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling Stage II")


enrichplot_Inflammatory_response_III <- enrichplot::gseaplot2(gseaHallmark_res.Stage_III, 
                      geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE", 
                      title = "Inflammatory response III")

enrichplot_Interferon_alpha_response_III <- enrichplot::gseaplot2(gseaHallmark_res.Stage_III, 
                      geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                      title = "Interfron alpha response III")

enrichplot_Interferon_gamma_response_III <- enrichplot::gseaplot2(gseaHallmark_res.Stage_III, 
                      geneSetID = "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                      title = "Interfron gamma response III")

enrichplot_TNFA_signaling_III <- enrichplot::gseaplot2(gseaHallmark_res.Stage_III, 
                      geneSetID = "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                      title = "TNFA signaling III")


# add line 
enrichplot_Inflammatory_response_Tumor[[1]] <- enrichplot_Inflammatory_response_Tumor[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_Interferon_alpha_response_Tumor[[1]] <- enrichplot_Interferon_alpha_response_Tumor[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_Interferon_gamma_response_Tumor[[1]] <- enrichplot_Interferon_gamma_response_Tumor[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_TNFA_signaling_Tumor[[1]] <- enrichplot_TNFA_signaling_Tumor[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")

enrichplot_Inflammatory_response_II[[1]] <- enrichplot_Inflammatory_response_II[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_Interferon_alpha_response_II[[1]] <- enrichplot_Interferon_alpha_response_II[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_Interferon_gamma_response_II[[1]] <- enrichplot_Interferon_gamma_response_II[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_TNFA_signaling_II[[1]] <- enrichplot_TNFA_signaling_II[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")

enrichplot_Inflammatory_response_III[[1]] <- enrichplot_Inflammatory_response_III[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_Interferon_alpha_response_III[[1]] <- enrichplot_Interferon_alpha_response_III[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_Interferon_gamma_response_III[[1]] <- enrichplot_Interferon_gamma_response_III[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")
enrichplot_TNFA_signaling_III[[1]] <- enrichplot_TNFA_signaling_III[[1]] + geom_hline(yintercept = 0, linetype = "solid", color = "black")

# show
enrichplot_Inflammatory_response_Tumor
enrichplot_Interferon_alpha_response_Tumor
enrichplot_Interferon_gamma_response_Tumor
enrichplot_TNFA_signaling_Tumor

enrichplot_Inflammatory_response_II
enrichplot_Interferon_alpha_response_II
enrichplot_Interferon_gamma_response_II
enrichplot_TNFA_signaling_II

enrichplot_Inflammatory_response_III
enrichplot_Interferon_alpha_response_III
enrichplot_Interferon_gamma_response_III
enrichplot_TNFA_signaling_III

```

* **Result comparison (prediction #2)**: All four pathways are an integral part of the immune system. When
looking at the enrichment score graph for the tumor_vs_non_tumor, we can
see that the genes are differentially suppressed in the tumor
microenvironment. In this case, we say that the tumor is "cold" meaning,
it suppresses immune activity.

We won't speak too much about the enrichment score of the stage
comparisons, but it is worth noting serveral things:

-   While the inflammatory response pathway is supressed in the tumor
    samples compared to the normal samples. We actually see activation
    in stage II and stage III compared to stage I.
-   Stage III has supressed the Interfron ALPHA response pathway more
    than stage II, while the Interfron GAMMA response pathway plot looks
    relatively similar between both comparisons
-   TNFA signaling pathway is just a tad more supressed in Stage III
    than in Stage II

### Tumor Microenvironment Deconvoluution

**Prediction #3: Immune Cell Supression**

-   **Prediction** Seeing as the 4 immune pathways we chose are
    supressed around the tumor compared to the normal tissue, we would
    assume that all if not some of the immune cells connected to that
    pathway would be supressed. Inflammatory Response: "Macrophages",
    "Neutrophils", "T-Cells", "B-Cells", "NK-Cells", "D-Cells" Interfron
    Alpha Response: "NK-Cells", "B-Cells", "T-Cells", "D-Cells"
    Interfron Gamma Response: "Macrophages", "T Helper Cells", "CD8+
    Cytotoxic T-Cells", "NK-Cells", "B-Cells", "D-Cells" TNFA signaling:
    "Macrophages", "T-Cells", "B-Cells", "NK-Cells", "Fibroblasts",
    "Keratinocytes"

But we predict that the Immune Cell Suppression by the different stages
of the cancer will not really be present. Because we have seen no major
contribution of the cancer stage to variability, and we saw that no
immune pathways were really suppressed (compared to Stage I cancer), and
they were even activated, that would tell us that the immune cell
enrichment will be a total mess.

-   **Biological reasoning**: The cells listed above are connected to
    the pathways we analyzed. And cancer varies a lot between people, so
    the stages will not really affect the immune cell enrichment, and it
    should be varied a lot.

```{r}
# first we get the normalized  counts from the differential expression
normalized_counts_matrix.LIHC_shortLetterCode <- counts(LIHC_shortLetterCode.dds, normalized = TRUE)
normalized_counts_matrix.LIHC_Stage <- counts(LIHC_Stage.dds, normalized = TRUE)

sum(is.na(rownames(normalized_counts_matrix.LIHC_shortLetterCode)))
sum(is.na(rownames(normalized_counts_matrix.LIHC_Stage)))

# remove version number
rownames(normalized_counts_matrix.LIHC_shortLetterCode) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC_shortLetterCode))

rownames(normalized_counts_matrix.LIHC_Stage) <- gsub("\\..*", "", rownames(normalized_counts_matrix.LIHC_Stage))

# make symbol (important to do twice because they dont have excatly the same data)
gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC_shortLetterCode),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC_shortLetterCode) <- gene_symbols[rownames(normalized_counts_matrix.LIHC_shortLetterCode)]

gene_symbols <- mapIds(org.Hs.eg.db,
                        keys = rownames(normalized_counts_matrix.LIHC_Stage),
                        column = "SYMBOL",
                        keytype = "ENSEMBL",
                        multiVals = "first")

rownames(normalized_counts_matrix.LIHC_Stage) <- gene_symbols[rownames(normalized_counts_matrix.LIHC_Stage)]

# no error! that's a relief. lets check for NA
sum(is.na(rownames(normalized_counts_matrix.LIHC_shortLetterCode)))
sum(is.na(rownames(normalized_counts_matrix.LIHC_Stage)))

# thats a lot of na, so lets get them out
clean_normalized_counts_matrix.LIHC_shortLetterCode <- normalized_counts_matrix.LIHC_shortLetterCode[!is.na(rownames(normalized_counts_matrix.LIHC_shortLetterCode)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC_shortLetterCode)))

clean_normalized_counts_matrix.LIHC_Stage <- normalized_counts_matrix.LIHC_Stage[!is.na(rownames(normalized_counts_matrix.LIHC_Stage)),]
sum(is.na(rownames(clean_normalized_counts_matrix.LIHC_Stage)))
```

**run xCell2**

```{r}
# start the actual analysis

# load immune reference
data(ImmuneCompendium.xCell2Ref, package = "xCell2")

# run analysis
xCell_immune_res.LIHC_shortLetterCode <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC_shortLetterCode,
                               xcell2object = ImmuneCompendium.xCell2Ref)

xCell_immune_res.LIHC_Stage <- xCell2Analysis(mix = clean_normalized_counts_matrix.LIHC_Stage,
                               xcell2object = ImmuneCompendium.xCell2Ref)
```

**visualize using heatmaps and cutoffs**

```{r fig.width=8, fig.height=8}
# lets visualize the results

# we need an "annotation col" for some reason
annotation_col <- data.frame(Condition = Bio.LIHC_shortLetterCode.clean_metadata_df$shortLetterCode)
rownames(annotation_col) <- rownames(Bio.LIHC_shortLetterCode.clean_metadata_df)

# order the samples by sample type
ordered_for_heatmap <- rownames(Bio.LIHC_shortLetterCode.clean_metadata_df)[order(Bio.LIHC_shortLetterCode.clean_metadata_df$shortLetterCode)]
ordered_xCell2_res <- xCell_immune_res.LIHC_shortLetterCode[, ordered_for_heatmap]
pheatmap(ordered_xCell2_res, 
         annotation_col = annotation_col,
         scale = "row",
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         breaks = seq(-2, 2, length.out = 100),
         cutree_rows = 5,
         main = "xCell2 cell type heatmap")

annotation_col <- data.frame(Condition = Bio.LIHC_Stage.clean_metadata_df$simplified_stage)
rownames(annotation_col) <- rownames(Bio.LIHC_Stage.clean_metadata_df)

# order the samples by sample type
ordered_for_heatmap <- rownames(Bio.LIHC_Stage.clean_metadata_df)[order(Bio.LIHC_Stage.clean_metadata_df$simplified_stage)]
ordered_xCell2_res <- xCell_immune_res.LIHC_Stage[, ordered_for_heatmap]
xCell_Stage_plot <- pheatmap(ordered_xCell2_res, 
         annotation_col = annotation_col,
         scale = "row",
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         breaks = seq(-3, 3, length.out = 100),
         cutree_rows = 3,
         main = "xCell2 cell type heatmap")



```
* **Result comparison (prediction #3)**:We see a clear immune deactivation in the tumor samples compared to the non tumor samples, when looking at: - granulocyte - natural killer
cell - monocyte - neutrophil - plasmacytoid dendritic cell - non
classical monocyte - eosinophil - classical nonocyte - basophil and more
with names too long. by looking at the hierarcial clustering and cutoff,
we can see that all supressed immune cells are closely related to each
other. and by asking gemini (don't know what else i can use), these are
all cells that are a big part of the "innate immune system" who's job is
to be the "first line of defense" when there is some type of attack.
When the Innate Immune System is activated for too long, the Adaptive
Immune System comes in. But with the Innate Immune System being active
in normal samples and not tumor ones could point maybe towards an immune
response, but not where the tumor is.

### 2.5 Results:

**Biological Integration**: Although these findings seem bleak, the innate immune system being the one suppressed would mean that treating that supression is a promising candidate for treating the cancer (Oncol Lett. 2020, 'Clinical value of innate immune cells in HCC' section, para. 4).
While the immunotherapy treatment is dependant on the cancer stage and BCLC staging (Lu, C., Rong, D., Zhang, B., Zheng, W., Wang, X., Chen, Z., & Liu, S. (2024). Current perspectives on the immune microenvironment in hepatocellular carcinoma: From pathogenesis to immunotherapy.), with the combination of the non correlated effect of the cancer stage on the immune system, a better solution might come that treats a wider amount of HCC cases of different stages.

------------------------------------------------------------------------

## 3. Secondary analysis: scRNA-Seq

### 3.1 Prediction #2

**Prediction**:

**Biological reasoning**:

### 2.3 Decision checkpoint #2:

**The decision**:

**Why we chose it**:

**Alternative considered**:

**Why we rejected the alternative**:

### 3.2 Code:

```{r harmony}
# load 1st dataset
sc_data_1 <- ReadMtx(
  mtx = "Data/scRNA_GSE140228/matrix.mtx",
  cells = "Data/scRNA_GSE140228/barcodes.tsv",
  features = "Data/scRNA_GSE140228/features.tsv",
  skip.feature = 1
)
seu_1 <- CreateSeuratObject(counts = sc_data_1, project = "GSE140228", min.cells = 3, min.features = 200)

# load 2nd dataset
file_path_2 <- "Data/scRNA_GSE149614/GSE149614_HCC.scRNAseq.S71915.count.txt"
sc_data_2_raw <- fread(file_path_2, data.table = FALSE, select = 1:3001)
rownames(sc_data_2_raw) <- sc_data_2_raw[, 1]
sc_data_2_clean <- sc_data_2_raw[, -1]
sc_data_2_sparse <- as(as.matrix(sc_data_2_clean), "dgCMatrix")

seu_2 <- CreateSeuratObject(counts = sc_data_2_sparse, project = "GSE149614", min.cells = 3, min.features = 200)

# merge and filter out low quality cells
seu_merged <- merge(seu_1, y = seu_2, add.cell.ids = c("DS1", "DS2"), project = "HCC_Integrated")
seu_merged[["percent.mt"]] <- PercentageFeatureSet(seu_merged, pattern = "^MT-")
seu_merged <- subset(seu_merged, subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mt < 10)

# pipeline and Harmony batch correction
seu_merged <- NormalizeData(seu_merged) %>% 
              FindVariableFeatures() %>% 
              ScaleData() %>% 
              RunPCA() %>% 
              RunHarmony(group.by.vars = "orig.ident")

# clustering and UMAP
seu_merged <- FindNeighbors(seu_merged, reduction = "harmony", dims = 1:15)
seu_merged <- FindClusters(seu_merged, resolution = 0.5)
seu_merged <- RunUMAP(seu_merged, reduction = "harmony", dims = 1:15)

# visual check plots
DimPlot(seu_merged, reduction = "umap", label = TRUE) + ggtitle("Cluster Map")
FeaturePlot(seu_merged, features = "CD8A") + ggtitle("CD8+ T-cell Localization")
```

We examined the UMAP and CD8A FeaturePlot to localize the T-cell
compartment, and identified clusters 0, 1, 6, 14, 17 as our target
populations.

### 3.2 Decision checkpoint #3: cluster characterization method

**The decision**: We chose to use logistic regression with dataset
integration.

**Why we chose it**: Although we integrated our datasets using Harmony,
Harmony only corrects the low dimensional embedding (PCA and UMAP) to
allow for shared clustering. It does not modify the underlying gene
expression matrix. By using logistic regression with latent.vars =
"orig.ident", we can explicitly account for and neutralize the batch
effects (technical variation between the two studies) during the actual
marker discovery. This ensures our markers are biologically driven by
cell type and not artifacts of different laboratories.

**Alternative considered**: Wilcoxon rank-sum test.

**Why we rejected the alternative**: The Wilcoxon test is a
non-parametric test that compares ranks but cannot incorporate
covariates. Using it on merged datasets without accounting for batch
effects often leads to "false positive" markers that represent dataset
differences rather than true cell-type biology.

### 3.3 Code:

```{r lr}
target_clusters <- c(0, 1, 6, 14, 17)

# join all layers
seu_merged <- JoinLayers(seu_merged)

# extract our T-cell group markers
t_markers <- FindMarkers(seu_merged, 
                         ident.1 = target_clusters, 
                         test.use = "LR", 
                         latent.vars = "orig.ident", 
                         only.pos = TRUE, 
                         min.pct = 0.25)

# keep only the significant genes
sig_genes <- rownames(t_markers[t_markers$p_val_adj < 0.05 & t_markers$avg_log2FC > 0.5, ])

# GO
ego_results <- enrichGO(gene          = sig_genes,
                        OrgDb         = org.Hs.eg.db,
                        keyType       = 'SYMBOL',
                        ont           = "BP", 
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.05)

dotplot(ego_results, showCategory = 15, label_format = 40) + 
  theme_bw() + 
  scale_color_gradient(low = "red", high = "blue") +
  ggtitle("Functional Enrichment: CD8+ T-cell Compartment") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

Now, we will examine if our target T-cells actively co-express classical
exhaustion markers. We will also test if this overall exhaustion
signature is significantly stronger in the tumor environment compared to
normal tissue:

```{r results}
exhaustion_list <- list(c("PDCD1", "LAG3", "HAVCR2", "TIGIT", "CTLA4", "TOX", "LAYN"))

seu_merged <- AddModuleScore(
  object = seu_merged,
  features = exhaustion_list,
  name = "T_Cell_Exhaustion_Score"
)

exhaustion_markers <- c("CD8A", "PDCD1", "LAG3", "HAVCR2", "TIGIT", "TOX")

p1 <- DotPlot(seu_merged, features = exhaustion_markers, idents = target_clusters) +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdYlBu"))) +
  labs(title = "Marker Expression in T-cell Clusters", x = "Genes", y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- FeaturePlot(seu_merged, features = "T_Cell_Exhaustion_Score1") +
  scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu"))) +
  ggtitle("Global T-Cell Exhaustion Score")

meta_ds1 <- read.delim("Data/scRNA_GSE140228/GSE140228_UMI_counts_Droplet_cellinfo.tsv", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds1) <- paste0("DS1_", meta_ds1$Barcode)
condition_ds1 <- meta_ds1[, "Tissue", drop = FALSE]
colnames(condition_ds1) <- "Condition"

meta_ds2 <- read.delim("Data/scRNA_GSE149614/GSE149614_HCC.metadata.updated.txt", header = TRUE, stringsAsFactors = FALSE)
rownames(meta_ds2) <- paste0("DS2_", meta_ds2$Cell)
condition_ds2 <- meta_ds2[, "site", drop = FALSE]
colnames(condition_ds2) <- "Condition"

combined_meta <- rbind(condition_ds1, condition_ds2)
seu_merged <- AddMetaData(seu_merged, metadata = combined_meta)

seu_merged$Condition[seu_merged$Condition == "Adjacent"] <- "Normal"

seu_filtered <- subset(seu_merged, subset = Condition %in% c("Tumor", "Normal"))

p3 <- VlnPlot(seu_filtered, features = "T_Cell_Exhaustion_Score1", 
              idents = target_clusters, group.by = "Condition", pt.size = 0) +
  ggtitle("Exhaustion Score: Normal vs Tumor") +
  stat_compare_means(method = "wilcox.test")

print(p1)
print(p2)
print(p3)
```

### 3.4 Results:

**Result comparison (prediction #2)**:

**Biological Integration**:
